### 2016-05-05 less-than-proof.tex
pushing master branch from iu

\title{Martin-L\"of Type Theory: Example proof}
\author{Carl Factora}
\date{\today}

\documentclass[12pt]{article}
% The following packages are needed because unicode
% is translated (using the next set of packages) to
% latex commands. You may need more packages if you
% use more unicode characters:
\usepackage{amssymb}
\usepackage{bbm}
\usepackage[greek,english]{babel}
% This handles the translation of unicode to latex:
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}

% Some characters that are not automatically defined
% (you figure out by the latex compilation errors you get),
% and you need to define:
\DeclareUnicodeCharacter{8988}{\ensuremath{\ulcorner}}
\DeclareUnicodeCharacter{8989}{\ensuremath{\urcorner}}
\DeclareUnicodeCharacter{8803}{\ensuremath{\overline{\equiv}}}
\DeclareUnicodeCharacter{120738}{\ensuremath{\Sigma}}
\DeclareUnicodeCharacter{955}{\ensuremath{\lambda}}

% Add more as you need them (shouldn’t happen often).
% Using “\newenvironment” to redefine verbatim to
% be called “code” doesn’t always work properly. 
% You can more reliably use:
\usepackage{fancyvrb}

\DefineVerbatimEnvironment
  {code}{Verbatim}
  {} % Add fancy options here if you like.

\begin{document}
\maketitle

\section*{Preliminary definitions}

\begin{code}
  open import Data.Nat using (ℕ; suc)
  open import Relation.Binary.PropositionalEquality
    using (_≡_; refl; cong; sym; trans)
  open import Data.Product
  open import Data.Empty

  -- Natural Number Recursion and Induction principle
  recℕ : (C : Set) → C → (ℕ → C → C) → ℕ → C
  recℕ C c f 0 = c
  recℕ C c f (suc n) = f n (recℕ C c f n)

  indℕ : (C : ℕ → Set) → C 0 → ((n : ℕ) → C n → C (suc n)) →
         (n : ℕ) → C n
  indℕ C c f zero = c
  indℕ C c f (suc n) = f n (indℕ C c f n)

  -- Addition
  add : ℕ → ℕ → ℕ
  add = recℕ (ℕ → ℕ) (λ n → n) (λ m r n → suc (r n))
\end{code}
We also define the following propositions:
\begin{code}
  ¬ : Set → Set
  ¬ A = A → ⊥

  _≤_ : (i j : ℕ) → Set
  i ≤ j = 𝞢[ k ∈ ℕ ] (add i k ≡ j)

  _<_ : (i j : ℕ) → Set
  i < j = (i ≤ j) × ¬ (i ≡ j)

  0≠si : (i : ℕ) → ¬ (0 ≡ suc i)
  0≠si i = λ ()

\end{code}
Here, we define {\tt ≤} and {\tt <} as:
\begin{itemize}
\item {\tt i ≤ j} is provable if and only if there exists a {\tt k} such that
  {\tt i + k = j}.
\item {\tt i < j} is provable if and only if we can provide a proof that
  {\tt i ≤ j} and that {\tt i ≠ j}.
\end{itemize}
Thus, both propositions are proven by constructing by pairs. The proof for
{\tt i ≤ j} must be a dependent pair since its corresponding proposition
uses a specific natural number (viz. {\tt k}).

{\tt 0≠si} is a proof that 0 is never equal to the successor of a natural
number. Here, we write {\tt λ ()} since the second assumption of the proof,
{\tt 0 ≡ suc i}, allows us to simply appeal to the fact that it is absurd to say
that 0 is equal to any natural number {\tt + 1}, by definition of the natural
numbers and the fact that Agda assumes that terms that are constructed
differently (i.e., 0 and {\tt suc} for natural numbers) can never be equal.

\section*{The Proof}
Using the above, we can prove the following:
\begin{code}
  0< : (i : ℕ) → (0 < suc i)
  0< = ?
\end{code}
The proof of this proposition requires that we prove that 0 is less than all
natural numbers. We do this using induction. More specifically, this proof
requires us to perform induction on a natural number, {\tt i}. To do
this, we use the {\tt indℕ} principle defined above. For each step of our proof,
we also provide the corresponding proof environment (viz. {\tt Proof Environment})
that represents the given proof obligation for each hole (viz. {\tt \{ \}}).
\begin{code}
0< : (i : ℕ) → (0 < suc i)
0< = indℕ { }0 { }1 { }2

{---Proof Environment
  ?0 : ℕ → Set
  ?1 : 0 < suc 0
  ?2 : (n : ℕ) → 0 < suc n → 0 < suc (suc n)
-}
\end{code}
The first of these obligations (i.e., {\tt ?0}) requires us to declare the
proposition we are wanting to prove. In general, this simply mirrors the proofs
type definition.
\begin{code}
0< : (i : ℕ) → (0 < suc i)
0< = indℕ (λ i → 0 < suc i) { }1 { }2

{---Proof Environment
  ?1 : 0 < suc 0
  ?2 : (n : ℕ) → 0 < suc n → 0 < suc (suc n)
-}
\end{code}
Our next obligation requires us to prove the proposition in the case that
\mbox{{\tt i = 0}} (i.e., we must show that {\tt 0 < suc 0}). Given our
definition of {\tt <}, the proof this by constructing a pair, containing proofs
of \mbox{{\tt 0 ≤ suc 0}} and \mbox{{\tt ¬ (0 ≡ suc 0)}}.

\begin{code}
0< : (i : ℕ) → (0 < suc i)
0< = indℕ (λ i → 0 < suc i)
          ({ }0 , { }1)
          { }2

{---Proof Environment
  ?0 : 0 ≤ suc 0
  ?1 : ¬ (0 ≡ suc 0)
  ?2 : (n : ℕ) → 0 < suc n → 0 < suc (suc n)
-}
\end{code}
To prove {\tt ?0}, we must construct a dependent pair that contains a natural
number {\tt k}, such that {\tt 0 + k = 1}. Thus, we construct a pair containing
the natural number 1 and as well as the proof that {\tt 1 = 1} (e.g. {\tt refl}).
To prove {\tt ?1}, we use our proof of {\tt 0≠si} on 0 to prove that
{\tt ¬ (0 ≡ suc 0)}.
\begin{code}
0< : (i : ℕ) → (0 < suc i)
0< = indℕ (λ i → 0 < suc i)
          ((1 , refl) , 0≠si 0)
          { }2

{---Proof Environment
  ?2 : (n : ℕ) → 0 < suc n → 0 < suc (suc n)
-}
\end{code}
To prove the final proposition, we must prove that for all natural numbers, {\tt n},
with the assumption that 0 is less than {\tt suc n}, we can show that 0 is less
than {\tt suc (suc n)} (i.e., {\tt n + 2}). 
\begin{code}
0< : (i : ℕ) → (0 < suc i)
0< = indℕ (λ i → 0 < suc i)
          ((1 , refl) , 0≠si 0)
          (λ n ih → { }0)

{---Proof Environment
  ?0 : 0 < suc (suc n)
-}
\end{code}
In this case, our inductive hypothesis, {\tt ih}, is actually not necessary,
since we can simply employ the same strategy as in the earlier proof for the
base-case.
\begin{code}
  0< : (i : ℕ) → (0 < suc i)
  0< = indℕ (λ i → 0 < suc i)
            ((1 , refl) , 0≠si 0)
            (λ n _ → ((suc (suc n)) , refl) , 0≠si (suc n))

\end{code}
\end{document}

### 2016-05-05 paper.tex
pushing master branch from iu

\title{Dependent Types and Martin-L\"of Type Theory: An Introduction}
\author{Carl Factora}
\date{\today}

\documentclass[12pt]{article}
% The following packages are needed because unicode
% is translated (using the next set of packages) to
% latex commands. You may need more packages if you
% use more unicode characters:
\usepackage{amssymb}
\usepackage{bbm}
\usepackage[greek,english]{babel}
% This handles the translation of unicode to latex:
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage{tikz}
\usetikzlibrary{matrix}
\usetikzlibrary{arrows}

% Some characters that are not automatically defined
% (you figure out by the latex compilation errors you get),
% and you need to define:
\DeclareUnicodeCharacter{8988}{\ensuremath{\ulcorner}}
\DeclareUnicodeCharacter{8989}{\ensuremath{\urcorner}}
\DeclareUnicodeCharacter{8803}{\ensuremath{\overline{\equiv}}}
\DeclareUnicodeCharacter{120738}{\ensuremath{\Sigma}}
\DeclareUnicodeCharacter{955}{\ensuremath{\lambda}}

% Add more as you need them (shouldn’t happen often).
% Using “\newenvironment” to redefine verbatim to
% be called “code” doesn’t always work properly. 
% You can more reliably use:
\usepackage{fancyvrb}

\DefineVerbatimEnvironment
  {code}{Verbatim}
  {} % Add fancy options here if you like.

\begin{document}
\maketitle

\begin{abstract}
  Who knows.
\end{abstract}

\section{Introduction}
Introduce the contents of the paper, notation, talk about Agda.

\section{Dependent Types}
In this section, we introduce the concept of dependent types and their role in
type theoretic systems. At their core, dependent types can be thought of simply
as an \textit{interplay of types and terms}. To understand this relationship,
it helps to think of the ways in which types and terms can \textit{depend} on
each other. Altogether, there are four different kinds of dependencies:

\begin{enumerate}
\item terms depending on terms
\item terms depending on types
\item types depending on types
\item types depending on terms
\end{enumerate}
These dependecies, however, should not be confused with an actual dependent type.
In fact, only (2) and (4) correspond to expressions that are typed (necessarily)
with a dependent type (we explain later why (1) and (3) are not typed with
dependent types).

What follows is a brief description of each form of dependency and the
corresponding \(\lambda\)-calculus that is associated with it. With this, we
introduce the (fully-dependent) system of the \textit{Calculus of Constructions}
(\(\lambda C\)). Here, the word \textit{fully} signifies that \(\lambda C\)
features all four forms of dependency mentioned above.

\subsection*{Simply-Typed (STLC)}
\subsubsection*{Role: Base System}
The STLC is not a dependently typed system. In fact, the term {\em dependent type}
is drawn in constrast with the term {\em simple type}, where the STLC derives
its namesake. A simple type is either a {\em primitive type}, which is the set
of all types ``pre-defined'' within the system, or a {\em function type} between
other simple types. Thus, simple types have the following grammar:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
t || t → t
\end{code}
\end{minipage}
\end{center}
The terms of the STLC are simply the terms of the \textit{untyped \(\lambda\)-calculus},
with the exception of terms that use \textit{self-application} (i.e., Y
combinator, Omega, etc.). This has the benefit of removing all \textit{infinite}
calculations from the system but also removing all recursive computations. Thus,
the STLC is not \textit{Turing complete} and is thus the appropriate foundation
for deriving a type theoretic system. It must first, however, be extended with
more expressive types, as the only form of dependency present in the STLC is (1),
terms depending on terms (i.e. {\em term level} \(\lambda\)-abstraction).

The key difference between the STLC and \(\lambda C\) are the places in which
terms and types are allowed to be. In the STLC, terms and types are entirely
seperate entities in that terms only exist on {\em term level} and types
on {\em type level}. Thus, as we mentioned earlier, this puts a constraint
on the system to only allow \(\lambda\)-abstraction to happen at term level
(viz. definition of simple type). For example, the \textit{constant function}
over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC looks like:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
const : ℕ → 𝔹 → ℕ
const = λ n b → n
\end{code}
\end{minipage}
\end{center}
Given that {\tt const} is of type {\tt ℕ → 𝔹 → ℕ}, the term that inhabits this
type is {\tt λ n b → n}. In this example, we can see that terms and types are
treated differently and are only related in terms of {\em inhabitance}. This
relation is reason for the limited expressivity of the STLC and is also the
method in which the STLC can be extended into \(\lambda C\).

\subsection*{Second-Order (\(\lambda2\))}
\subsubsection*{Role: Addition of polymorphic functions}

Another name for \(\lambda2\) is System F. For our purposes, we treat
\(\lambda2\) as simply an extension of the STLC. This system is adds
{\em polymorphic} terms to the STLC (viz. Dependency (2)). For example,
let's take the definition of {\tt const} above and change it to define
{\tt polyConst}:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
polyConst : (A B : Set) → A → B → A
polyConst A B = λ a b → a
-- Works for any type A and B
\end{code}
\end{minipage}
\end{center}

Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
are types. These two types (viz {\tt A} and {\tt B}) are the corresponding
types of the third and fourth arguments of {\tt polyConst}. To clarify, we can
instantiate {\tt polyConst} with \(\mathbb{N}\) and \(\mathbb{B}\) resulting
in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us

$${\tt const} = {\tt polyConst(ℕ,𝔹)}.$$
Thus, adding polymorphism to the STLC has the effect of extending the notion
of type and term. With \(\lambda2\), we are now allowed to pass types as arguments
at term level (viz. we pass the actual types \(\mathbb{N}\) and \(\mathbb{B}\)).
It is, however, necessary to constrain this expressivity with a more expressive
type (e.g. {\tt (x : A) → ...}). Types that appear in this fashion are
called \(\Pi\)-types. This leads us to a new grammar for types ({\tt t}):

\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
t || t → t || ∀ (x : Set) → t
\end{code}

\end{minipage}
\begin{minipage}{0.65\textwidth}
\begin{small}
In general, the \(\forall\) symbol is used to express \(\Pi\)-types
\end{small}
\end{minipage}
\end{center}

Briefly, \(\Pi\)-types are a form of dependent type that are inhabited by
polymorphic functions. In cases like {\tt polyConst}, \(\Pi\)-types are
necessary to signify the treament of a polymorphic function and to prevent
the appearance of free variables within type expressions. For example,
if we had typed {\tt polyConst} as {\tt A → B → A}, then {\tt A} and {\tt B} are
free variables in that they, by themselves, have no {\em meaning}. This is why
{\tt polyConst} is first passed two types to instantiate the meanings of {\tt A}
and {\tt B}. In a sense, a \(\Pi\)-type binds values for types just as
\(\lambda\)-abstraction binds values for terms.

\(\Pi\)-types are actually significantly more expressive than the ones in
\(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
over other types (e.g. {\tt t} in {\tt ∀ (x : Set) → t}). In a later system, we
allow \(\Pi\)-types to be parameterized over terms (e.g. {\tt ∀ (x : t) → t}).

So far, we've seen how terms can depend on other terms (STLC) and how terms can
depend on types (term level polymoprhism). In the following system, we introduce
a small, yet important dependency: types depending on types.

\subsection*{Higher-order (\(\lambda\underline{\omega}\))}
\subsubsection*{Role: Addition of type constructors}

To clarify, \(\lambda\underline{\omega}\) is generally not a standalone system,
instead, it is generally treated as an ``add-on'' system. Later, we model the
relationship of each system to \(\lambda C\). For the sake of time, we treat
\(\lambda\underline{\omega}\) as an extension of \(\lambda2\), giving us the
following transition:

\begin{center}
\begin{minipage}{0.4\textwidth}
STLC \(\rightarrow\) \(\lambda2\) \(\rightarrow\) \(\lambda\underline{\omega}\) \(\rightarrow\) \(\lambda\omega\)
\end{minipage}
\end{center}

Thus, the system we are introducing is \(\lambda\omega\), which, unlike
\(\lambda\underline{\omega}\), features a more diverse grammar of types (i.e.,
\(\lambda\underline{\omega}\) does not have \(\Pi\)-types). Given its \(\lambda2\)
component, another name for this system is System F\(\omega\). Before we introduce
the new grammar for types, we first introduce the concept of {\em levels}.

We are already familiar with the first two levels, terms and types, which, from
the STLC, are still related in terms of inhabitance. To meet the constraints of
the STLC when introducing type constructors to our system, we must also provide
the type of types. For our purposes, we assume an infinite universe of types, as
is common practice in deriving type theoretic systems to avoid certain paradoxes.
For example, let's first assume that natural numbers are primitive within our
system. We then have the following for the natural number 42:

\begin{center}
\begin{minipage}{0.4\textwidth}
\begin{code}
42 : ℕ : Set : Set₁ ...
\end{code}
\end{minipage}
\end{center}

Another name for {\tt Set} is a {\em kind}. For our puroses, we use the
more familiar set-theoretic equivalent reading. Reading the above example is rather
straightforward: 42 is of type \(\mathbb{N}\) which is of type {\tt Set}
which of type {\tt Set₁} and so on. There is, however, an important distinction
for the first two levels: 42 is NOT of type {\tt Set}, while \(\mathbb{N}\) is
both of type {\tt Set} and {\tt Set₁}. Terms are therefore treated uniquely,
since they only exist as elements of types. While important, this topic is a bit
out of the scope of an introductory paper, however, we briefly discuss this
in a later section.

We can now revise the definition of types ({\tt t}) to include all instances of
{\tt Set} in our infinite universe. We notate this as {\tt Set} below:

\begin{center}
\begin{minipage}{0.65\textwidth}
\begin{code}
t || t → t || ∀ (x : Set) → t || Set
\end{code}
\end{minipage}
\end{center}

The curious effect of doing this allows us to construct expressions of type
{\tt t → t}, where either {\tt t} can have subexpressions containing {\tt Set}.
Let's take a familiar definition of the list data structure:

\begin{center}
\begin{minipage}{0.65\textwidth}
\begin{code}
data List (A : Set) : Set where
  Nil  : List A
  _::_ : A → List A → List A
\end{code}
\end{minipage}
\end{center}

We discuss {\tt List} more in depth in a later section. For now, we simply ask
{\em what is the type of {\tt List}?} To answer this question, we can simply
look at the first line of the definition above (viz. {\tt List (A : Set) : Set}).
From this line, we can see that {\tt List} is a type parameterized over any type
{\tt A}. Another way to think about this statement is to think of {\tt List} as
a form of polymorphic function of type {\tt Set → Set}, as is similar to other
{\em parameterized types}, that can be applied at type level (thus the name
{\em type constructor}). This is to insure that we cannot have create type
expressions that fail to instantiate {\tt List} (i.e., {\tt List → A}).

From this we can see that the simple addition of {\tt Set} into our grammar of
types has great effect on the expressivity of our system. There is only one
other extension necessary to consider our system fully-dependently typed, which,
similar to our treatment of \(\lambda\underline{\omega}\) is also deceptively
simple.

\subsection*{Predicates (\(\lambda P\))}
\subsubsection*{Role: Encoding of Propositional logic}

To review, if we compile all the previously mentioned systems above, we have
the system \(\lambda\omega\) (System F\(\omega\)). Our final addition, the
system of \(\lambda P\), allows us to construct {\em type families} which are
meant to directly correspond to propositions in higher-order logic.

Let's take a simple example. Let's define a proposition {\tt isEven} that takes
a natural number. Another way to say this is that {\tt isEven} is a predicate
over natural numbers. When writing propositions, we write them as the equivalent
type definition (similar to {\tt List}). Consequently, the proposition
{\tt isEven} is an example of a dependent type, since its {\em meaning} depends
on the natural number passed to it.

\begin{center}
\begin{minipage}{0.7\textwidth}
\begin{code}
data _isEven : ℕ → Set where
  ZEven : 0 isEven
  NEven : ∀ n → n isEven → (Suc (Suc n)) isEven
\end{code}
\end{minipage}
\end{center}

This is the last form of dependency of the four we mentioned earlier. Here,
we've defined two ways in which {\tt n isEven} for all natural numbers
({\tt n}) can be proven ``true'', which is done via its two type constructors
{\tt ZEven} and {\tt NEven}, corresponding to the two constructors of
\(\mathbb{N}\). {\tt ZEven} corresponds to the proof that 0 is an even number.
{\tt NEven} then defines the property that if a given natural number ({\tt n}) is
even (viz. {\tt n isEven}), then {\tt n + 2} is even (viz. {\tt (Suc (Suc n))
isEven}). To clarify how this type can be used, let's prove that 4 is even. We
start by providing the appropriate type definition and also providing a name for
our proof (viz. {\tt 4isEven}):
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
\end{code}
\end{minipage}
\end{center}
We must then show that 4 is indeed even using a combination of {\tt ZEven} and
{\tt NEven}. From the definition of {\tt isEven}, we know that to prove 4 is
even, we must know that 2 is even.  Consequently, to prove 2 is even, we must
know that 0 is even. Fortunately, we have a proof that 0 is even: {\tt ZEven}.
Thus, the proof that 2 is even is {\tt NEven 0 ZEven}, and finally the proof
that 4 is even is:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
4isEven : 4 isEven
4isEven = NEven 2 (NEven 0 ZEven)
\end{code}
\end{minipage}
\end{center}

This method of proof is adopted from {\em Intuitionistic Logic}. In this form of
logic, ``truth'' corresponds to the idea of {\em type inhabitability}. This
makes sense since we use define types to correspond to logical propositions
(e.g. {\tt isEven}). ``Falsity'' then naturally corresponds to un-inhabitability
of a given type. This can be seen if we try to prove that 5 is even using
{\tt isEven}. The proof seems to get ``stuck.''

While it might seem intuitive and more effective to instead try to prove that 5
is {\em not} even, this, however, also doesn't work. This is due to the
treatment of negation in Intuitionistic Logic. To prove that 5 is not even is
to prove that \(\neg (5\) {\tt isEven}), which is equivalent to constructing a
function of type:
\begin{center}
\begin{minipage}{0.2\textwidth}
\begin{code}
5 isEven → ⊥
\end{code}
\end{minipage}
\end{center}
That is, given a proof that {\tt 5 isEven}, construct \(\perp\), the empty type.
This is not possible since we cannot ever construct a proof that 5 is even. We
discuss the empty type later in Section 3.

\subsection*{Calculus of Constructions and Remarks}
To conclude our discussion over dependent types, we present the final grammar for
types after extending \(\lambda\omega\) with \(\lambda P\):
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
t || t → t || ∀ (x : t) → t || Set
\end{code}
\end{minipage}
\end{center}
It should come with no surprise that the change is minor. We simply have lifted
the restriction on \(\Pi\)-types and allowed them to be parameterized over both
terms and types.

In the process of introducing dependent types, we have shown the necessary
changes and extensions on the STLC to achieve the expressivity necessary to
function as a foundation for mathematics. To clarify this process and to show
the relationship between each of the systems we have previously mentioned, we
include a diagram of the {\em Lambda Cube}:
\begin{center}
  \begin{tikzpicture}[
      back line/.style={densely dotted},
      cross line/.style={preaction={draw=white, -,
          line width=6pt}}]
    \matrix (m) [matrix of math nodes,
      row sep=1em, column sep=.75em,
      text height=1ex,
      text depth=0.5ex]{
      & \lambda\omega & & \lambda C \\
      \lambda 2 & & \lambda P2 \\
      & \lambda\underline\omega & & \lambda P\underline\omega \\
      \lambda\!\!\rightarrow  & & \lambda P \\
    };
    \path[-]
    (m-1-2) edge (m-1-4)
    edge (m-2-1)
    edge [back line] (m-3-2)
    (m-1-4) edge (m-3-4)
    edge (m-2-3)
    (m-2-1) edge [cross line] (m-2-3)
    edge (m-4-1)
    (m-3-2) edge [back line] (m-3-4)
    edge [back line] (m-4-1)
    (m-4-1) edge (m-4-3)
    (m-3-4) edge (m-4-3)
    (m-2-3) edge [cross line] (m-4-3);
  \end{tikzpicture}
\end{center}
Reading the cube requires us to start on the edge marked \(\lambda\rightarrow\),
which corresponds to the STLC, the base system. From here, we can either move
inwards, upwards, or rightwards, which corresponds to the extension to the
other systems marked on the respective edges. Using only these three movements,
we eventually reach \(\lambda C\), leaving it in the {\em most powerful}
position on the cube.

\section{Martin-L\"of Type Theory}
%% Curry-Howard?
In this section, we introduce the foundational concepts of Martin-L\"of Type
Theory. We discuss recursion, induction, and as well several primitive type
encodings (e.g. from propositional logic). We recommend the reader to have a
running proof assistant while reading this section to help fortify (better word)
several concepts.

\subsection*{Principles}
Recursion and induction are one of the powerful tools available in a
fully-dependently typed system. Using recursion and induction principles to
define functions and proofs, we are guaranteed termination and totality. This is
because, in general, recursion and induction principles are defined under certain
known intuitions on the given type that they are meant to act on and thus also
follow a certain pattern.
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
recT : (C : Set) → C₀ ... → T → C
indT : (C : T → Set) → (C₀ : ...) ... → (t : T) → C t
\end{code}
\end{minipage}

\begin{minipage}{0.7\textwidth}
\begin{small}

The recursion and induction principles for some type {\tt T}.
\end{small}
\end{minipage}
\end{center}
The first argument given to these principles is sometimes called the {\em motive}.
This simply means that {\tt C} is the type we are going to inhabit. For recursion
principles, this means the given output of the function we are defining, and for
induction principles, this can be read as the proposition we are trying to prove.
The terms that follow (viz. {\tt C₀}) correspond to the constructors of the given
type, which is how these principles guarantee totality. Finally, both principles
take a term of the type that are acting on (viz. {\tt T} and {\tt (t : T)}). The
only difference is that the induction principle introduces a type family, as
opposed to non-parameterized type in the recursion principle. We discuss this in
Section 4.

In this section, we discuss the recursion and induction principles, {\tt rec}
and {\tt ind}, for each {\em primitive type}: \(\perp\), \(\top\), +,
\(\times\), and \(\Sigma\). The symbols + and \(\times\) signify disjunction
and conjunction and, thus, should not be confused with addition and
multiplication. For the latter, we use the names {\tt plus} and {\tt times}.

\subsection*{Primitive Type Encodings}
%% Bottom, Top, Disjunction, Conjunction, Sigma

\section{Example Types and Sample Proofs}
Describe the section.

\subsection*{Booleans}
\subsection*{Natural Numbers}
\subsection*{List/Orderings}
\subsection*{Identity Types}

\section{Discussion}
Say a few things about the paper.

\subsection*{Extension to Homotopy Type Theory}

\end{document}

### 2016-05-05 lambda-c.tex
pushing old version

%-----------------------------------------------------------------------------
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%-----------------------------------------------------------------------------
\documentclass[9pt,preprint,nocopyrightspace,computermodern]{sigplanconf} %% ,preprint,nocopyrightspace

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{graphicx, color, xcolor}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{fontspec}
\setmonofont[Scale=MatchLowercase]{Courier}
\usepackage{epigraph}
\usepackage{pmboxdraw}
\usepackage{textcomp}
\usepackage{acronym}
\usepackage{float}
\usepackage{balance}
\usepackage{amssymb}

%% mine
\usepackage{mathpartir}
\usepackage{titlesec}
\usepackage{proof}
\usepackage{tabu}
\usepackage{array}
\usepackage{tikz}
\usepackage{caption}
\usetikzlibrary{matrix}
\usetikzlibrary{arrows}

%%%%-------------------------------------------------------------------------------------
%% Old format stuff

%% \documentclass[18pt,twocolumn] {article}

%% \usepackage{amssymb}
%% \usepackage{amsmath}
%% \usepackage{mathpartir}
%% \usepackage{titlesec}
%% \usepackage{proof}
%% \usepackage{tabu}
%% \usepackage{array}
%% \usepackage{tikz}
%% \usetikzlibrary{matrix}

%% \titlespacing\section{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}
%% \titlespacing\subsection{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}
%% \titlespacing\subsubsection{0pt}{12pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt}

%% % Math-mode symbol & verbatim
%% \def\W#1#2{$#1{#2}$ &\tt\string#1\string{#2\string}}
%% \def\X#1{$#1$ &\tt\string#1}
%% \def\Y#1{$\big#1$ &\tt\string#1}
%% \def\Z#1{\tt\string#1}

%% % A non-floating table environment.
%% \makeatletter
%% \renewenvironment{table}%
%%    {\vskip\intextsep\parskip\z@
%%     \vbox\bgroup\centering\def\@captype{table}}%
%%    {\egroup\vskip\intextsep}
%% \makeatother

%% % All the tables are \label'ed in case this document ever gets some
%% % explanatory text written, however there are no \refs as yet. To save
%% % LaTeX-ing the file twice we go:
%% \renewcommand{\label}[1]{}
%%%%-------------------------------------------------------------------------------------
\begin{document}
\bibliographystyle{plainnat}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}
\setlength{\epigraphrule}{0pt}
%%\copyrightyear{2014} 
%% \copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
%% \copyrightdata{TBA} 
%% \doi{TBA}

% \exclusivelicense                % ACM gets exclusive license to publish, 
% you retain copyright

%%\permissiontopublish             % ACM gets nonexclusive license to publish
% (paid open-access papers, 
% short abstracts)

%% \preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Calculus of Constructions: {\em Type Theory in Pieces}}
\authorinfo{Carl Factora \quad Daniel P. Friedman}{August 19, 2015}{}{}

\maketitle

\begin{abstract}
  The Calculus of Constructions, \(\lambda C\), is the most powerful and expressive
  \(\lambda\)-Calculus. \(\lambda C\) does not have the drawbacks that its components,
  \(\lambda\!\!\rightarrow, \lambda2, \lambda\underline\omega\) and \(\lambda P\),
  have on their own, but instead retains many of their positive attributes, allowing
  it to more accurately and successfully embody the ideas of \textit{Type Theory} and
  serve effectively as a foundation for both logic and mathematics.
\end{abstract}

\section*{Preliminaries}
The reader is assumed to have some familiarity with the \textit{untyped lambda calculus},
here denoted by the symbol \textbf{\(\lambda\)}. The section covering \(\lambda\) serves
as a basic overview of the more important aspects of the calculus that are also present in
\(\lambda C\) and as well as the other calculi.

The reader is also assumed to have some familiarity with reading \textit{derivation rules}.
Briefly, each derivation rule is read left to right, top to bottom. The statement(s) above
the line are the \textit{premises} for the expression below the line. For type theoretic
rules, the expression \(\Gamma\vdash\! X\) means that from the context \(\Gamma\), a list
of premisses, we can derive \(X\). For logical derivation rules, we include the general
\textit{natural deduction} rules of \textit{intro} and \textit{elim}. Intro rules define
how a logical expression is created from context, and elim rules define how logical expressions
are handled and \textit{reduced}.

Should the reader wish to read more in depth on the concepts we offer here, we refer them
to the book \mbox{``Type Theory and Formal Proof''} \cite{rng} by Nederpelt and \mbox{Geuvers},
from which we have modified several examples and offer our own explanations and insights.

For the sake of time and space, each section begins with notable \textit{Pros} and
\textit{Cons} associated with the given calculus. We also include \textit{extension maps}
for every section that describes the given calculus' distance from \(\lambda C\).
In general, \(\lambda\) isn't actually a part of \(\lambda C\). We include
\(\lambda\) in the extension maps since \(\lambda\!\!\rightarrow\) is an extension
of \(\lambda\).

\section*{1. Untyped Lambda Calculus (\(\lambda\))}
\begin{tikzpicture}
  \draw [dashed] (0,0) -- (1.5,0);
  \draw [dashed] (1.5,0) -- (3,0);
  \draw [dashed] (3,0) -- (4.5,0);
  \draw [dashed] (4.5,0) -- (6,0);
  \draw [ultra thick] (0,-.1) node[below]{\(\lambda\)} -- (0,0.1);
  \draw [thick] (1.5,-.1) node[below]{?} -- (1.5,0.1);
  \draw [thick] (3,-.1) node[below]{?} -- (3,0.1);
  \draw [thick] (4.5,-.1) node[below]{?} -- (4.5,0.1);
  \draw [thick] (6,-.1) node[below]{\(\lambda C\)} -- (6,0.1);
\end{tikzpicture}
\begin{flushleft}
  \textbf{Pros}: simple, \textit{Turing complete}
  \par
  \textbf{Cons}: inclusion of \textit{infinite calculations}
\end{flushleft}
The untyped lambda calculus \((\lambda)\) is a simple foundation for all other calculi.
As we have mentioned earlier, \(\lambda\) is not necessarily a component of \(\lambda C\),
despite its position on the extension map above.

\(\lambda\) can be thought of as a language comprised of three building blocks, which
are called \mbox{\textit{\(\Lambda\)-terms}~\cite{bar1}.}
\mbox{\(\Lambda\)-terms \((\Lambda)\)} are comprised of:
\begin{itemize}
\item (\(V\)) Variables (\(x,y,z ...\))
\item (\(L\)) Abstractions (\(\lambda V.\,\Lambda\))
\item Applications (\(\Lambda_0\,\Lambda_1\))
\end{itemize}
In the above definition, the third bullet shows that the two \(\Lambda\)-terms
in an \textit{application} could be different. The second bullet states
that the term immediately following the \(\lambda\) symbol in
an \textit{abstraction} must be one of the \textit{variables} of \((V)\).

\(\lambda\) has three derivation rules, for variables, abstractions and
applications. This leaves the calculus exceptionally simple, as the
entire idea of the calculus is focused around the relationship between the three
rules and their respective \mbox{\(\Lambda\)-terms.}

An important aspect of \(\lambda\) is the notion of \textit{\(\beta\)-equivalence}.
\(\beta\)-equivalence states that two \(\Lambda\)-terms are \(\beta\)-equivalent
if there exists a way to reduce one term via \((app)\) into the second.
We give the definition of \((app)\) and an example of \(\beta\)-equivalence below.

\begin{center}
  \infer[{(app)}]
        {\Gamma\vdash\,B[x := \Lambda]}
        {\Gamma\vdash\lambda x\,.\,B & \Gamma\vdash\Lambda}\par
\end{center}

\begin{itemize}
\item \(((\lambda x\,.\,x)y)\,=_\beta\,y\), by \textit{one} use of \((app)\)
\item In general, parentheses are left out of \(\Lambda\)-terms when it is clear how
  the term should be evaluated. The application \((\lambda x\,.\,\lambda y\,.\,x)\,a\,b\)
  is shorthand for \((((\lambda x\,.\,(\lambda y\,.\,x))\,a)\,b)\), because
  applications are \textit{right associative}.
\item In a similar vein, \((\lambda x,y\,.\,x)a\,b\), although not used here,
  represents the same application given above.
\item \((\lambda x\,.\,\lambda y\,.\,x)\,a\,b,\,((\lambda y\,.\,x)[x := a])b\),
  and \(a\) are \mbox{\(\beta\)-equivalent} terms.
\end{itemize}
The notation [\(x\) := \(a\)] signifies \textit{substitution}, in which all free
occurrences of the variable \textit{x} within the term it is placed beside are
replaced by \textit{a}. In the above example, \(x\) occurs free in \((\lambda y\,.\,x)\),
thus substitution results in the term \((\lambda y\,.\,a)\). The idea of substitution
is a key point within all lambda calculi, as it embodies the way in which
expressions are  \textit{evaluated}. \mbox{\(\beta\)-equivalence} is important
later on when its use is extended to the \textit{\(\beta\)-equivalence of types}.

This simple system is expressive enough to be \textit{Turing complete}, meaning that
it can represent every possible calculation. This also leads, however, to the
calculus' greatest drawback: \textit{infinite calculations}. In other words,
\(\lambda\) is too expressive and not \textit{restrictive} enough when we are
defining legal terms. This is why an extension of \(\lambda\) is necessary to prevent
such calculations from happening. A natural way to remedy this is to add
\textit{typing restrictions}, which are added in \(\lambda\!\!\rightarrow\).

\section*{2. Simply-Typed (\(\lambda\!\!\rightarrow\))}
\begin{tikzpicture}
  \draw [-] (0,0) -- (1.5,0);
  \draw [dashed] (1.5,0) -- (3,0);
  \draw [dashed] (3,0) -- (4.5,0);
  \draw [dashed] (4.5,0) -- (6,0);
  \draw [thick] (0,-.1) node[below]{\(\lambda\)} -- (0,0.1);
  \draw [ultra thick] (1.5,-.1) node[below]{\(\lambda\!\!\rightarrow\)} -- (1.5,0.1);
  \draw [thick] (3,-.1) node[below]{?} -- (3,0.1);
  \draw [thick] (4.5,-.1) node[below]{?} -- (4.5,0.1);
  \draw [thick] (6,-.1) node[below]{\(\lambda C\)} -- (6,0.1);
\end{tikzpicture}
\begin{flushleft}
  \textbf{Pros}: No \textit{infinite calculations}
  \par
  \textbf{Cons}: Restricted expressivity
\end{flushleft}
The simply-typed \(\lambda\)-calculus \((\lambda\!\!\!\rightarrow)\) can be
summarized rather succintly. It is simply an extension of \(\lambda\) with
\textit{simple types} \((T)\). A simple type \cite{church} is defined as follows:
\begin{itemize}
\item (\(V_T\)) Type variable \((\alpha,\beta,\gamma\,...)\)
\item Arrow type \((T\,\rightarrow\,T)\)
\end{itemize}
This definition states that the only types available in \(\lambda\!\!\rightarrow\)
are made up of \textit{type variables} and \textit{arrow types}, hence the name
\(\lambda\!\!\rightarrow\). The general rule of \(\lambda\!\!\rightarrow\)
requires that \(\Lambda\)-terms be typed. The notation for representing
\(\Lambda\)-terms is then slightly changed from \(\lambda\). Examples of this
notation are given below.
\begin{enumerate}
  \item \(M : \alpha\), for an arbitrary \(\Lambda\)-term \(M\)
  \item \((\lambda (x : \alpha)\,.\,N)\,:\,\alpha\rightarrow\gamma\), provided that \(N : \gamma\).
    In general, this term will be represented as \(\lambda x : \alpha\,.\,N\,:\,\alpha\rightarrow\gamma\).
  %% \item \((\lambda x : \alpha\,.\,N)M : \gamma\rightarrow_\beta N[x := M] : \gamma\)
\end{enumerate}
The above notation is called \textit{Church typing}, as opposed to ``Curry typing'',
and is what we use throughout, since it is the notation most commonly used in Type Theory.
For \((1)\), the notation is read as ``the term \(M\) of type \(\alpha\).''
In \((2)\), the notation is read as ``the term \(N\) abstracted over by \(x\)
of type \(\alpha\) is of type \(\alpha\rightarrow\gamma\).'' In general, the body of
an abstraction (viz. \(N\)) is not typed explicitly and is instead inferred from the
type of the entire abstraction. %% Example \((3)\) represents an application of
%% \((2)\) on \((1)\), resulting in an expression of type \(\gamma\). After
%% \(\beta\)-reduction, the resulting term retains the type \(\gamma\).
Thus, in \(\lambda\!\!\rightarrow\), the ``:'' symbol separates a
\(\Lambda\)-term and its respective simple type.

The primary motivation for \(\lambda\!\!\rightarrow\) is the removal of infinite
calculations. When a system does not have infinite calculations, it is described
as \textit{strongly normalizing}. \(\lambda\!\!\rightarrow\) achieves this through a
modification of its \((app)\) rule. With these modifications, \(\beta\)-reduction
can only happen when a given term is \textit{well-typed}.
\begin{center}
  \infer[{(app)}]
        {\Gamma\vdash\,MN\,:\,\beta}
        {\Gamma\vdash\, M\,:\,\alpha\rightarrow\beta & \Gamma\vdash\, N\,:\,\alpha}
\end{center}
To summarize the above rule, application is only possible if the value passed to
an abstraction corresponds to its input type. Here, the abstraction
\(\lambda x : \alpha\,.\,N\) requires an input of type \(\alpha\). To see this
more clearly, we include an example of a \textit{legal} and \textit{illegal}
application below.
\begin{itemize}
\item \((\lambda x : \alpha\,.\,N\,:\,\alpha\rightarrow\gamma)(Y\,:\,\alpha)\)
  \(\rightarrow_\beta\) \(N[x := Y] : \gamma\)
\item \((\lambda x : \alpha\,.\,N\,:\,\alpha\rightarrow\gamma)(M\,:\,\sigma)\)
  cannot be \(\beta\)-reduced as \mbox{\(\sigma\) \(\not =\) \(\alpha\).}
\end{itemize}
The symbol \(\rightarrow_\beta\) signifies a \textit{one step \(\beta\)-reduction},
meaning that the term to the right of the \(\rightarrow_\beta\) symbol is derived
from the term on the left after one use of \((app)\). From the above example, the
effects of incorrect inputs can be seen. Simply restricting the number of possible
inputs to a given function is surprisingly enough to prevent
\textit{infinite calculations} \cite{san}.

The problem with \(\lambda\!\!\rightarrow\) stems from the same attribute:
\textit{the types are too simple to hold more complicated information},
which, consequently, inhibits the expressivity of the calculus. In order to
remedy this, \(\lambda\!\!\rightarrow\) can be extended with more expressive types.
Once this is done, the resulting system retains the attribute of being strongly
normalizing while also being more expressive.

\section*{3. Second-Order (\(\lambda2\))}
\begin{tikzpicture}
  \draw [-] (0,0) -- (1.5,0);
  \draw [-] (1.5,0) -- (3,0);
  \draw [dashed] (3,0) -- (4.5,0);
  \draw [dashed] (4.5,0) -- (6,0);
  \draw [thick] (0,-.1) node[below]{\(\lambda\)} -- (0,0.1);
  \draw [thick] (1.5,-.1) node[below]{\(\lambda\!\!\rightarrow\)} -- (1.5,0.1);
  \draw [ultra thick] (3,-.1) node[below]{\(\lambda 2\)} -- (3,0.1);
  \draw [thick] (4.5,-.1) node[below]{?} -- (4.5,0.1);
  \draw [thick] (6,-.1) node[below]{\(\lambda C\)} -- (6,0.1);
\end{tikzpicture}
\begin{flushleft}
  \textbf{Pros}: \textit{terms depending on types} (\(\Pi\)types)
  \par
  \textbf{Cons}: Lacks other dependencies
\end{flushleft}
The second-order lambda calculus \((\lambda2\)) derives its namesake from the
manner in which it extends \(\lambda\!\!\rightarrow\). \(\lambda 2\) categorizes
its expressions on \textit{two levels}: terms making up the first level and types
the second. In Type Theory, there are \textit{four levels}. We describe the four
levels in the section on \(\lambda\underline\omega\) (Section 4).

\(\lambda 2\) focuses on the addition of \textit{\(\Pi\)types}. \(\Pi\)types can
be thought of as a way of accurately representing \textit{dependencies} between
types and terms. In a sense, the same idea is encapsulated with arrow types but is
only used when the given dependency does not refer to \textit{terms depending on terms}.

In fact, \textit{terms depending on terms} are already present in \(\lambda\!\!\rightarrow\)
and \(\lambda\) when defining abstractions. The terms \(\lambda x\,.\,N\) and
\(\lambda x:\alpha\,.\,N : \alpha\rightarrow\gamma\) for \(\lambda\) and
\(\lambda\!\!\rightarrow\), respectively, are examples of terms depending on terms,
which in both cases is \(N\) depending on \(x\).

Since \(\lambda2\) is an extension of \(\lambda\!\!\rightarrow\), we must
define the \textit{type of all types} before \(\Pi\)types can be added, since
\(\lambda\!\!\rightarrow\) requires that everything be typed. Here we use the asterisk symbol,
\(*\), to represent the type of all types. Thus, all type variables
(\(\alpha,\beta,\gamma ...\)) are of type \(*\).

Adding \textit{terms depending on types} to \(\lambda\!\!\rightarrow\) raises the level
of expressivity and results in the system \(\lambda 2\). To show what specifically
has been added and also, consequently, what a \textit{term depending on a type} looks like,
we present the following example below.
\begin{itemize}
\item In \(\lambda\), the \textit{identity} function is \(\lambda x\,.\,x\), which functions
  for all possible inputs (e.g. Naturals, Integers, Booleans)
\item In \(\lambda\!\!\rightarrow\), there are too many identity functions
  (one for each simple type), when there should only be one.
\item To fix this, a \textit{polymorphic} identity function is made possible with
  \(\lambda2\). A function is \textit{polymorphic} when it generalizes over types
  (i.e., it can be applied to terms of different types). 
\item The polymorphic identity function is \(\lambda\,\alpha\,:\,*\,.\,\lambda\,x\,:\,\alpha\,.\,x\)
\end{itemize}
This particular polymorphic identity function can effectively serve as \textit{the}
identity function for all possible inputs \cite{gir}. For example, the function application
\((\lambda\alpha:*\,.\,\lambda x:\alpha\,.\,x) nat\) serves as the identity function
for Naturals since it \(\beta\)-reduces to \(\lambda x:nat\,.\,x\).

Typing the polymorphic identity function requires a \(\Pi\)type. This is necessary
because all polymorphic functions represent a form of dependency, and thus their
types should do the same. Hence, the most appropriate type for the polymorphic identity
function is \(\Pi\alpha:*\,.\,\alpha\rightarrow\alpha\).

From this we can see that the notion of types has been extended with a new type:
\(\Pi\)type. With \(\Pi\)types, the notion of \textit{simple types} has been removed
and the new definition for \textit{types} \((T)\) is:
\begin{itemize}
\item \((V_T)\) Type variable \((\alpha,\beta,\gamma\,...)\)
\item Arrow Type \((T\,\rightarrow\,T)\)
\item \(\Pi\)type \((\Pi V_T:*\,.\,T)\)
\end{itemize}
Consequently, the definition for \(\Lambda\)-terms \((\Lambda)\) has also changed.
The new definition for \(\Lambda\)-terms \cite{gir2}, which now includes dependent
terms, is given below.
\begin{itemize}
\item \((V)\) Term variable \((x,y,z...)\)
\item \((L)\) Abstraction \((\lambda V:T\,.\,\Lambda)\)
\item Application \((\Lambda_0\,\Lambda_1)\)
\item \((L_\Pi)\) Universal Abstraction \((\lambda V_T:*\,.\,\Lambda)\)
\item Universal Application \((L_\Pi\,T)\)
\end{itemize}
Having two instances of abstraction and application seems a bit excessive, however,
with what is explained later, the two variations are essentially equivalent if
we treat arrow types as \textit{special cases} of \(\Pi\)types.

Even with the addition of \(\Pi\)types and the expressivity that they bring,
\(\lambda2\) retains the strongly normalizing attribute of \(\lambda\!\!\!\rightarrow\).
From here, we may add different notions of type and term dependency until we reach
the full expressivity of \(\lambda C\).
%% \(\lambda2\), however, still does not have enough expressivity to function as an
%% effective foundation for logic or mathematics. It is still missing different notions
%% of type and term dependency.

%% TO DO: Fix beginning to state that we are adding \(\lambda\underline\omega\)
%% to lambda 2, resulting in \(\lambda\omega\). Mention that in this section
%% we are essentially describing the extension, not the system as a whole, since
%% that system is simply a combination of the two, nothing more. Don't forget
%% to mention System Fw

%% Also, fix the ending to make a transition into Lambda P.
\section*{4. Higher-Order (\(\lambda\underline\omega\))}
\begin{tikzpicture}
  \draw [-] (0,0) -- (1.5,0);
  \draw [-] (1.5,0) -- (3,0);
  \draw [dashed] (3,0) -- (4.5,0);
  \draw [dashed] (4.5,0) -- (6,0);
  \draw [thick] (0,-.1) node[below]{\(\lambda\)} -- (0,0.1);
  \draw [thick] (1.5,-.1) node[below]{\(\lambda\!\!\rightarrow\)} -- (1.5,0.1);
  \draw [thick] (3,-.1) node[below]{\(\lambda 2\)} -- (3,0.1);
  \draw [ultra thick] (3.75,0.1) node[above]{\(+\lambda\underline\omega\)} -- (3.75,0.21);
  \draw [thick] (4.5,-.1) node[below]{\(\lambda\omega\)} -- (4.5,0.1);
  \draw [thick] (6,-.1) node[below]{\(\lambda C\)} -- (6,0.1);
\end{tikzpicture}
\begin{flushleft}
  \textbf{Pros}: \textit{types depending on types}
  \par  
  \textbf{Cons}: No \(\Pi\)types
\end{flushleft}
Another name for \(\lambda\underline\omega\) is the
\textit{higher order \(\lambda\)-calculus} \cite{gir}. This system,
when combined with \(\lambda 2\) results in the system \(F_\omega\) or,
as notated above, \(\lambda\omega\). In this section, we primarily
describe the smaller \(\lambda\underline\omega\), an extension
from \(\lambda\!\!\rightarrow\), and not the combination.

\(\lambda\underline\omega\) adds a powerful construct to \(\lambda\!\!\rightarrow\):
\textit{type constructors}. In popular dependently typed programming languages
(e.g. Idris, Epigram, Agda), a common example of a type constructor is \textit{List}.
In these languages, to instantiate an object of type \textit{List} requires an
additional type. This additional type functions as the type of the instantiated list.
It is clear here that \textit{List} also has a type, namely \(*\rightarrow *\). As
an example, when the type constructor \textit{List} is given the type \(\sigma\),
it returns a list of type \(\sigma\) (i.e., \(List_\sigma\) or (List \(\sigma\))).
\textit{Type constructors} are thus functions that return types.

Similar to the requirements of adding \(\Pi\)types in \(\lambda2\), to add the
notion of \textit{types depending on types} in \(\lambda\underline\omega\), the
resulting expressions must be properly typed. For this, we add the notion
of \textit{kinds} \((K)\). A kind is one of:
\begin{itemize}
\item \(*\)
\item \(K\rightarrow K\)
\end{itemize}
Thus, all type constructors can be typed using a kind. Consequently, kinds must
also be typed, and all kinds have the type \(\square\) (e.g. \(*\), \(*\rightarrow *\),
\((*\rightarrow *)\rightarrow *\) are of type \(\square\)).

It is now appropriate to introduce the \textit{four levels} of Type Theory and
explain the relationship that the levels have with each other. The four levels are:
\begin{enumerate}
\item Terms only (specifically \(\Lambda\)-terms)
\item Types and type constructors
\item Kinds only
\item \(\square\) only
\end{enumerate}
The relationship between the levels is somewhat subtle while at the same time
intuitive. All expressions can be typed by an expression of the following level.
Grasping the relationship between the levels is important for understanding
the derivation rules of \(\lambda\underline\omega\), \(\lambda P\) (Section 5)
and \(\lambda C\) (Section 6), since most of the key premises of their rules include
\textit{sorts}. A sort (\(s\)) is one of:
\begin{itemize}
\item \(*\)
\item \(\square\)
\end{itemize}
In general, a sort is used to type \textit{types} and the type of types, \(*\),
a requirement of \(\lambda\!\!\rightarrow\). It is also a key aspect in determining,
at any given point in a \textit{derivation}, which two levels we are working in.
To clarify, the following example details how the \((conv)\) rule of
\(\lambda\underline\omega\) relates with the \textit{\(\beta\)-equivalence of types}:
\begin{itemize}
\item Take the type constructor \(\lambda\alpha :*\,.\,\alpha\rightarrow\alpha\,:\,*\rightarrow *\).
\item This constructor requires a type of type \(*\), so any arbitrary type (say \(\gamma\)) works.
\item Then, \((\lambda\alpha :*\,.\,\alpha\rightarrow\alpha)\)\(\gamma\) is a type.
\item Let \(A : (\lambda\alpha :*\,.\,\alpha\rightarrow\alpha)\gamma\).
\item Intuitively, we would want to be able to \textit{reduce} this to say that \(A : \gamma\rightarrow\gamma\).
\end{itemize}
Enabling the last point above requires \textit{type convertibility}. This notion is
captured by the \((conv)\) rule given below.
\begin{center}
  \infer[{if B =_\beta B' \quad (conv)}]
        {\Gamma\vdash\,A\,:\,B'}
        {\Gamma\vdash\, A\,:\,B & \Gamma\vdash\, B'\,:\,s}
\end{center}
We can read the above rule as follows: \(A\) can be of type \(B\) and \(B'\) if
\(B'\) is a sort, with a side condition that \(B =_\beta B'\). Depending on the value
of the sort, \(s\), we can determine whether \(A\), \(B\) and \(B'\) are terms or types.
For example, if \(s\) is \(*\), then \(B\) and \(B'\) are types and \(A\) must be a term.
Similarly, if \(s\) is \(\square\), then \(B\) and \(B'\) are kinds and \(A\)  must be a type.
Since \(\lambda\underline\omega\) can have applications of type constructors in its
type expressions and all legal applications must have \(\beta\)-reductions, we use
type convertibility rather often.

Thus, \(\lambda\underline\omega\) adds type constructors to \(\lambda\!\!\rightarrow\),
but it no longer has the expressivity of \(\Pi\)types. Extending \(\lambda\underline\omega\)
with \(\Pi\)types, however, is simple and is thus often done in the creation of
\textit{hybrid} systems, an example of which is \(\lambda\omega\) (\(\lambda 2\) +
\(\lambda\underline\omega\)). From here, we are only missing one form of term
and type dependency (i.e., \textit{types depending on terms}) before we reach the
expressivity of \(\lambda C\). We take the time to describe this dependency and
extension in the following section.

%% Doing so results in an entirely different system called
%% \(\lambda\omega\) (without the underscore) or \(F\omega\) \cite{gir}.There is only
%% one kind of dependency left, namely that of \textit{types depending on terms}, which
%% is added in \(\lambda P\).

%% TO DO: Fix introduction. Explain that we are essentially in a ``crossroads'' between
%% the above system and Lambda C. Once we add the extension describe in this section,
%% then we reach Lambda C. Keep the fact that this is a system that pays an integeral
%% role for making up Lambda C
\section*{5. Predicate Lambda Calculus (\(\lambda P\))}
\begin{tikzpicture}
  \draw [-] (0,0) -- (1.5,0);
  \draw [-] (1.5,0) -- (3,0);
  \draw [-] (3,0) -- (4.5,0);
  \draw [dashed] (4.5,0) -- (6,0);
  \draw [thick] (0,-.1) node[below]{\(\lambda\)} -- (0,0.1);
  \draw [thick] (1.5,-.1) node[below]{\(\lambda\!\!\rightarrow\)} -- (1.5,0.1);
  \draw [thick] (3,-.1) node[below]{\(\lambda 2\)} -- (3,0.1);
  \draw [thick] (4.5,-.1) node[below]{\(\lambda\omega\)} -- (4.5,0.1);
  \draw [ultra thick] (5.25,0.1) node[above]{\(+\lambda P\)} -- (5.25,0.21);
  \draw [thick] (6,-.1) node[below]{\(\lambda C\)} -- (6,0.1);
\end{tikzpicture}
\begin{flushleft}
  \textbf{Pros}: \textit{types depending on terms}
  \par  
  \textbf{Cons}: Cannot encode \(\exists,\,\,\neg,\,\,\wedge,\,\, \text{and} \,\, \vee\)
\end{flushleft}
In this section, we set out to explain the final component of \(\lambda C\), namely
\(\lambda P\). \(\lambda P\) gives us \textit{types depending on terms}.

Out of all the components of \(\lambda C\) (Section 6), \(\lambda P\) plays one of
the more important roles in allowing \(\lambda C\) to serve as a possible foundation
for logic and mathematics. This is because \(\lambda P\) adds \textit{types depending on terms}
enabling it to encode \textit{minimal predicate logic} (described below), hence the
name \(\lambda P\). For our purposes, the overview of \(\lambda P\) describes how it does
this and why it is important.

\(\lambda P\) adds the notion of \textit{propositions} and \textit{predicates over types}.
To describe these entities and show what they look like in \(\lambda P\), the following example
details the type theoretic equivalent of the logical term:
\begin{center}
  \begin{tabu} {  X[c] | X[c] || X[c]  }
    \textbf{Inhabitants} & \textbf{Types} & \textbf{\(\lambda P\)} \\[0.05cm]
    \hline
    terms  & types & \(e : \alpha\) \\[0.05cm]
    \hline
    proofs & propositions & \(e : \rho\) \\[0.05cm]
  \end{tabu}\par\smallskip
  \textbf{Table 1:} PAT Interpretation
\end{center}
These entities correspond to the \textit{PAT-interpretation} of logic, a fundamental
aspect of Type Theory. PAT stands for \textit{propositions as types} and also
\textit{proofs as terms} \cite{how}. Together, they explain the correspondence between
\textit{Intuitionistic propositional logic} and Type Theory \cite{lof}.

In the more familiar Classical logic, every term can be evaluated to true or false.
In Intuitionistic Logic, the notion of truth corresponds to \textit{provability}.
Thus the ``truth'' of an expression \(B\) is determined by whether or not it can be proven
in a given context. In Type Theory, we represent this with the notion of \textit{inhabitation}.

We are already familiar with the notion of inhabitation, since the expression \(e : \rho\) can
also be read as \textit{the term e inhabits the type \(\rho\)} or \textit{e proves \(\rho\)}.
Returning to the idea of propositions as types, it is then clear why propositions \textit{are}
types, since they can be proven (\textit{inhabited}) by terms (proofs as terms).

Putting this all together, predicates then encapsulate the notion of
\textit{types depending on terms} in \(\lambda P\). For clarification, we give the
following example \cite{rng}:
\begin{itemize}
\item A predicate is of type \(\sigma\rightarrow *\), since it maps over an arbitrary type
  (here \(\sigma\)) and returns a proposition, which is a type \((*)\).
\item Let \(\rho_n\) be the proposition that \textit{n is a prime number}.
\item Propositions are sometimes notated as \(P(n)\). This notation states that the term
  \(n\) is allowed to appear \textit{free} in the term \(P\).
\item This notation is commonly mistaken for function invocation, so, for clarity, we use \(\rho_n\).
\item Then, \(\lambda n : nat\,.\,\rho_n\) is a predicate. Here, the type \(\rho_n\) depends
  on the value of the term \(n\).
\end{itemize}
Passing a Natural (say \(x\)) into \(\lambda n : nat\,.\,\rho_n\) would then return the
proposition \((\rho_x)\) that \(x\) is a prime number, which is only proven if an inhabitant
of type \(\rho_x\) can be found.

It is now appropriate to show the type theoretic encoding for \(\Rightarrow\) and \(\forall\),
which are the only connectives available in minimal predicate logic.%%  Below are the corresponding
%% \(\lambda P\) derivation rules that accurately model the behavior of the logical connectives of
%% minimal predicate logic.
\begin{center}
  \begin{tabu} {  X[c] | X[c]  }
  %% \begin{tabular}{ m{2.5cm} | m{2.5cm}  } 
    \textbf{Logic} & \textbf{Type Theory} \\[0.05cm]
    \hline
    \(\Rightarrow\)  & \(\rightarrow\)  \\[0.05cm]
    \hline
    \(A\Rightarrow B\)  & \(A\rightarrow B\)  \\[0.05cm]
    \hline
    \(\Rightarrow\)-intro  & \((abst)\)  \\[0.05cm]
    \hline
    \(\Rightarrow\)-elim  & \((app)\)  \\[0.05cm]
    \hline
    \hline
    \(\forall\)  & \(\Pi\)  \\[0.05cm]
    \hline
    \(\forall_ {x:\alpha} (\rho_x)\)  & \(\Pi x : \alpha\,.\,\rho_x\) \\[0.05cm]
    \hline
    \(\forall\)-intro  & \((abst)\)  \\[0.05cm]
    \hline
    \(\forall\)-elim  & \((app)\)  \\[0.05cm]
  \end{tabu}\par\smallskip
  \textbf{Table 2:} Type Theory \& Logic Correspondence
\end{center}
It seems strange that both the logical \textit{intro} and \textit{elim} rules for
\(\Rightarrow\) and \(\forall\) are modeled by the same derivation rule but are not
modeled by the same type theoretic connective. Returning to what we have mentioned in
the section for \(\lambda\underline\omega\), \(\lambda P\) treats \(\Pi\)types as a
special case for arrow types. Since \(\lambda P\) does not have the dependencies
mentioned in \(\lambda\underline\omega\) and \(\lambda 2\), it only needs to worry
about \textit{types depending on terms}. Clearly, \(\Pi\)types are \textit{only}
used when there is a dependency involved, which in \(\lambda P\) is present in terms
resembling \textit{predicates}. In arrow types, and consequently \(\Rightarrow\), the
expression to the right of \(\rightarrow\) can be thought of as \textit{independent}
from the expression on the left. To clarify this, the following example is given:
\begin{itemize}
\item \(\lambda x:\alpha\,.\,N\) can be modeled by an arrow type if \(x\)
  does not appear free in \(N\).
\item \(\lambda x:\alpha\,.\,\rho_x\) can only be modeled by a \(\Pi\)type since
  the term \(\rho_x\) depends on \(x\).
\end{itemize}
\(\lambda P\), however, has only one derivation rule for each \((abst)\) and \((app)\)
and the abstractions in these rules are always typed with a \(\Pi\)type. This means
that \(\Pi\)types can effectively model arrow types, since they hold more
information than \(\rightarrow\).

With the addition of \(\lambda P\), we have now seen all the various dependencies between
types and terms and the interesting constructs that they bring. We can now remove
the \textit{Cons} of all previously mentioned calculi by combining \(\lambda\!\!\rightarrow\),
\(\lambda 2\), \(\lambda\underline\omega\), and \(\lambda P\), resulting in \(\lambda C\).

\section*{6. Calculus of Constructions (\(\lambda C\))}
\begin{tikzpicture}
  \draw [-] (0,0) -- (1.5,0);
  \draw [-] (1.5,0) -- (3,0);
  \draw [-] (3,0) -- (4.5,0);
  \draw [-] (4.5,0) -- (6,0);
  \draw [thick] (0,-.1) node[below]{\(\lambda\)} -- (0,0.1);
  \draw [thick] (1.5,-.1) node[below]{\(\lambda\!\!\rightarrow\)} -- (1.5,0.1);
  \draw [thick] (3,-.1) node[below]{\(\lambda 2\)} -- (3,0.1);
  \draw [thick] (3,0.1) node[above]{\em Complete} -- (3,0.1);
  \draw [thick] (4.5,-.1) node[below]{\(\lambda\omega\)} -- (4.5,0.1);
  \draw [ultra thick] (6,-.1) node[below]{\(\lambda C\)} -- (6,0.1);
\end{tikzpicture}
\begin{flushleft}
  \textbf{Pros}: Able to fully encode \textit{Constructive Logic}
  \par  
  \textbf{Cons}: ?
\end{flushleft}
The Calculus of Constructions (\(\lambda C\)) is the culmination of the \(\lambda\)-calculi
we have mentioned and described. It retains the strongly normalizing attribute of
its components and their various forms of type and term dependency. With this power,
\(\lambda C\) can effectively serve as a foundation for logic and mathematics and
as well a \textit{typed programming language}.

The figure below is Barendregt's \(\lambda\)-cube \cite{bar2} and notates the
relationship between the calculi we mentioned earlier. The base of the cube
starts with \(\lambda\!\!\rightarrow\). From there, extending upwards results in
\(\lambda 2\), extending to a third dimension results in \(\lambda\underline\omega\),
and extending rightwards results in \(\lambda P\). From these other points, it is
also possible to extend to other calculi in the same manner. For example, \(\lambda P\)
can be extended with either \(\lambda2\) or \(\lambda\underline\omega\),
resulting in \(\lambda P2\) or \(\lambda P\underline\omega\) respectively.
This means that there actually are multiple ways to reach \(\lambda C\) from
\(\lambda\!\!\rightarrow\). Here, we first extended \(\lambda\!\!\rightarrow\)
first with \(\lambda2\), then added \(\lambda\underline\omega\) to reach
\(\lambda\omega\), then added \(\lambda P\) to reach \(\lambda C\). Looking again
at the cube, this leaves \(\lambda C\) in the position of the cube that
encompasses \textit{all} possible extensions from \(\lambda\!\!\rightarrow\).
\begin{center}
  \begin{tikzpicture}[
      back line/.style={densely dotted},
      cross line/.style={preaction={draw=white, -,
          line width=6pt}}]
    \matrix (m) [matrix of math nodes,
      row sep=1em, column sep=.75em,
      text height=1ex,
      text depth=0.5ex]{
      & \lambda\omega & & \lambda C \\
      \lambda 2 & & \lambda P2 \\
      & \lambda\underline\omega & & \lambda P\underline\omega \\
      \lambda\!\!\rightarrow  & & \lambda P \\
    };
    \path[-]
    (m-1-2) edge (m-1-4)
    edge (m-2-1)
    edge [back line] (m-3-2)
    (m-1-4) edge (m-3-4)
    edge (m-2-3)
    (m-2-1) edge [cross line] (m-2-3)
    edge (m-4-1)
    (m-3-2) edge [back line] (m-3-4)
    edge [back line] (m-4-1)
    (m-4-1) edge (m-4-3)
    (m-3-4) edge (m-4-3)
    (m-2-3) edge [cross line] (m-4-3);
  \end{tikzpicture}
\end{center}

\(\lambda C\) derives its name from Constructive Logic by its creator Thierry Coquand \cite{coq1}.
\(\lambda C\) is the \textit{only} \(\lambda\)-calculus that can fully encode the
logical connectives (\(C\)) of Intuitionistic Logic. It does this by combining
different forms of type and term dependencies from its components. The type of
these encodings are given below.
\begin{center}
  \begin{tabular}{ m{0.3cm} | m{7.4cm}  } 
    \textbf{C} & \textbf{\(\lambda C\) Type} \\[0.1cm]
    \hline      
    \(\perp\)  & \(\Pi\alpha : *\,.\,\alpha\)  \\[0.1cm]
    \hline
    \(\neg\)  & \(\lambda\alpha : *\,.\,(\alpha\rightarrow\perp)\)  \\[0.1cm]
    \hline
    \(\Rightarrow\)  & \(\lambda\alpha : *\,.\,\lambda\beta : *\,.\,\alpha\rightarrow\beta\)  \\[0.1cm]
    \hline
    \(\wedge\)  & \(\lambda\alpha : *\,.\,\lambda\beta : *\,.\,\Pi\gamma : *\,.\,(\alpha\rightarrow\beta\rightarrow\gamma)\rightarrow\gamma\)  \\[0.1cm]
    \hline
    \(\vee\)  & \(\lambda\alpha : *\,.\,\lambda\beta : *\,.\,\Pi\gamma : *\,.\,(\alpha\rightarrow\gamma)\rightarrow(\beta\rightarrow\gamma)\rightarrow\gamma\)  \\[0.1cm]
    \hline
    \(\forall\)  & \(\lambda\sigma : *\,.\,\lambda\rho : \sigma\rightarrow *\,.\,\Pi x : \sigma\,.\,\rho_x\)  \\[0.1cm]
    \hline
    \(\exists\)  & \(\lambda\sigma :*\,.\,\lambda\rho : \sigma\rightarrow *\,.\,\Pi\alpha : *\,.\,((\Pi x : \sigma\,.\,(\rho_x\rightarrow\alpha))\rightarrow\alpha\)  \\[0.1cm]
  \end{tabular}\par\smallskip
  \textbf{Table 3:} Connective Encodings
\end{center}
The types of these connectives are type constructors that return propositions
(types). Some of these type constructors can be encoded by a smaller subset
of \(\lambda C\) (i.e., \(\perp\) can be encoded by \(\lambda 2\)),
but some require a larger combination (i.e., \(\exists\) can only be encoded by
\(\lambda C\)). Below is a table containing the necessary system to encode each logical connective.
\begin{center}
  \begin{tabu} {  X[c] | X[c]  }
    \textbf{Connective} & \textbf{System Required} \\[0.05cm]
    \hline
    \(\perp\)  & \(\lambda2\)  \\[0.05cm]
    \hline
    \(\neg\)  &  \(\lambda\omega\) \\[0.05cm]
    \hline
    \(\Rightarrow\)  &  \(\lambda2\) or \(\lambda\underline\omega\) \\[0.05cm]
    \hline
    \(\wedge\)  &  \(\lambda\omega\) \\[0.05cm]
    \hline
    \(\vee\)  &   \(\lambda\omega\) \\[0.05cm]
    \hline
    \(\forall\)  & \(\lambda P2\) or \(\lambda P\underline\omega\) \\[0.05cm]
    \hline
    \(\exists\)  & \(\lambda C\)  \\[0.05cm]
  \end{tabu}\par\smallskip
  \textbf{Table 4:} System Requirements
\end{center}
The simplest way to explain how these encodings accurately model their respective
connective is to show how they are \textit{introduced} and \textit{eliminated}.
In \(\lambda C\), this is done solely by \((abst)\) and \((app)\), respectively.
Thus, the rules mentioned in the following sections are special cases of \((abst)\)
and \((app)\) modeled specfically for their respective connectives.

\subsubsection* {Negation (\(\neg\)) and Absurdity (\(\perp\))}
We describe \(\neg\) and \(\perp\) together because they are most commonly used
for one purpose: \textit{ex falso}. \textit{Ex falso} is a logical term that means
``from absurdity comes anything one wants.'' In Type Theory, this is equivalent to
``if \(\perp\) is inhabited, then everything is inhabited''. This behavior is modeled
by \(\perp\!\!-elim\). To find an inhabitant of \(\perp\), we must first derive a term
of type \(\perp\) via \(\perp\!\!-intro\). Both of these rules are given below.
\begin{center}
  \begin{tabu}{ X[c]  X[c] }
    \infer[{(\perp\!\! -intro)}]{NM:\,\perp}{M:A & N:A \rightarrow\perp} &
    \infer[{(\perp\!\! -elim)}]{NA:A}{N:\,\perp}  \\
  \end{tabu}
\end{center}
The type theoretic reading for \(\perp\!\! -intro\) is
\textit{if we can derive a term of type A and a term of type \(A\rightarrow\perp\), we can derive \(\perp\)}.
For clarification, the example below shows how these rules are used in a general
type theoretic setting.
\begin{itemize}
\item Assume that \(m\) is of type \(A\)
\item Assume that \(n\) is of type \(\neg A\)
\item By definition, \(n\) is also of type \(A\rightarrow\perp\)
\item Then, \(n\,m : \,\perp\)
\item Say we want to inhabit the type \(\gamma\)
\item From the type of \(\perp\), \(n\,m\) is also of type \(\Pi\alpha:*\,.\,\alpha\)
\item Then, \(n\,m\,\gamma : \gamma\) (i.e. \(n\,m\,\gamma\) proves \(\gamma\))
\end{itemize}

\subsubsection* {Implication (\(\Rightarrow\))}
There is a rather clear one-to-one correspondence between \(\Rightarrow\) and \(\rightarrow\).
The encoding for \(\Rightarrow\) we mentioned earlier is simply the polymorphic version,
as are all others the others in Table 3. This means that it takes two types and
creates an arrow type that maps the first type into the second. Like \(\lambda P\),
the arrow type of \(\lambda C\) is actually a \(\Pi\)type, so \((abst)\) and \((app)\)
are sufficient enough to model the proper behavior of \(\Rightarrow\).

\subsubsection* {Conjunction (\(\wedge\)) and Disjunction (\(\vee\))}
The encodings for \(\wedge\) and \(\vee\) correspond to their \textit{second-order encodings}.
The second order encoding is necessary, because the first order encoding is only
compatible with Classical logic. Since Intuitionistic logic is essentially a
\textit{subset} of Classical logic, its rules must be more general \cite{lof}.
In a simple example, conjunction corresponds to a typed \textit{pair} and
disjunction corresponds to a \textit{disjoint union}.
\begin{itemize}
\item Assume \(x:\alpha\), \(y:\beta\)
\item Assume \(z\) is a pair comprised of \(x\) and \(y\)
\item Then, \(z:(\alpha\times\beta)\), respresenting a conjunction of \(x\) and \(y\).
  Here, we can deconstruct \(z\) and get \(x:\alpha\) and \(y:\beta\).
\item For disjunction, deconstructing an expression of type \((\alpha + \beta)\)
  requires \textit{case analysis} or pattern matching. When we deconstruct
  such a term, we get either a term of type \(\alpha\) or a term of type \(\beta\).
\end{itemize}

%% Hence,
%% the encodings in \(\lambda C\) must model the more general versions of the \(\wedge\)
%% and \(\vee\) rules of Intuitionistic logic, included below.
%% \begin{center}
%%   \begin{tabu}{ X[c] }
%%     \infer[{(\wedge\!\! -\!intro)}]{\Pi C:*\,.\,(A\rightarrow B\rightarrow C)\rightarrow C}{A & B} \\
%%     \infer[{(\wedge\!\! -\!elim_1)}]{A}{\Pi C:*\,.\,(A\rightarrow B\rightarrow C)\rightarrow C} \\
%%     \infer[{(\wedge\!\! -\!elim_2)}]{B}{\Pi C:*\,.\,(A\rightarrow B\rightarrow C)\rightarrow C} \\
%%     \infer[{(\vee\!\! -\!intro_1)}]{\Pi C:*\,.\,(A\rightarrow C)\rightarrow(B\rightarrow C)\rightarrow C}{A} \\
%%     \infer[{(\vee\!\! -\!intro_2)}]{\Pi C:*\,.\,(A\rightarrow C)\rightarrow(B\rightarrow C)\rightarrow C}{B} \\
%%     \infer[{{\footnotesize (\vee\!\! -\!elim\!)}}]{C}{\Pi C:*\,.\,(A\rightarrow C)\rightarrow(B\rightarrow C)\rightarrow C & A\rightarrow C & B\rightarrow C} \\
%%   \end{tabu}
%% \end{center}
\subsubsection* {Quantifiers (\(\forall\) and \(\exists\))}
Similar to \(\Rightarrow\) and \(\rightarrow\), \(\forall\) is modeled accurately by
\(\Pi\). Like \(\lambda P\), the \((abst)\) and \((app)\) rules of \(\lambda C\) are
already perfect models for the rules of \(\forall\).

\(\exists\), however, is quite a bit more complicated. To clarify, the following
example details \(\exists\!\!-\!\!elim\) rule.
\begin{itemize}
\item Assume \(\sigma:*\),\(\,\,\rho:\sigma\rightarrow *\), and \(A:*\)
\item Assume \(y:\Pi\alpha:*\,.\,((\Pi x:\sigma\,.\,(\rho_x\rightarrow\alpha))\rightarrow\alpha)\)
\item Assume \(z:\Pi x:\sigma\,.\,(\rho_x\rightarrow A)\)
\item Then, \(y\,A:(\Pi x:\sigma\,.\,(\rho_x\rightarrow A))\rightarrow A)\)
\item \(y\,A\) is an abstraction: \(\lambda u:(\Pi x:\sigma\,.\,(\rho_x\rightarrow A))\,.\,A\)
\item Then, \(y\,A\,z:A\)
\end{itemize}
In the above example, \textit{y} is an encoding of the existential quantifier,
\(\exists\), and \(z\) is an encoding for the universal quantifier, \(\forall\).
To derive \(y\), we use \(\exists\!\!-\!\!intro\). To do so, we need a type,
a predicate over said type, an inhabitant of said type, and a proof of said predicate.
\begin{itemize}
\item Assume \(\sigma:*\),\(\,\,\rho:\sigma\rightarrow *\),\(\,\,a:\sigma\), and \(u:\rho_a\)
\item With these assumptions, we may simply abstract over \(a\) and \(u\) to
  create an inhabitant of \(\exists\).
\item Then, \(\lambda\alpha:*\,.\,\lambda v:(\Pi x:\sigma\,.\,(\rho_x\rightarrow\alpha))\,.\,v\,a\,u :\Pi\alpha:*\,.\,((\Pi x:\sigma\,.\,(\rho_x\rightarrow\alpha))\rightarrow\alpha)\)
\end{itemize}
This encoding of the existential quantifier is only possible in \(\lambda C\),
since it requires all forms of term and type dependency. This power and expressivity
found in \(\lambda C\) is the reason why it is commonly used as a foundation for
proof assistants. \(\lambda C\), however, is also commonly extended with
\textit{inductive} types, allowing it to prove properties on recursive data structures.

\subsubsection*{Inductive Constructions}
The extension of \(\lambda C\) with inductive types allows an encoding of recursive
data structures and functions, while retaining the strongly normalizing attribute of
\(\lambda\!\!\rightarrow\). This system is called the
\textit{Calculus of Inductive Constructions} \cite{bert}.

To show the power of this system, we present a proof of the \textit{double-reverse theorem}
in Idris. The double-reverse theorem states that the reverse of the reverse of
an arbitrary list, \(l\), is equal to \(l\). Below is the theorem we use for our proof.

{\small
\begin{verbatim}
drev_thm : (xs : List a) -> rev (rev xs) = xs
\end{verbatim}
}

In our type expressions, we exclude any implicit \(\Pi\)types that Idris can infer.
Specifically, we exclude the \(\Pi\)type {\small\tt \{a:Type\}}, since it is
clear in our type expressions that {\small\tt a} is a type.

To show the usage and general behavior of proof assistants, we use Idris'
\textit{interactive prover} to prove the double-reverse theorem. Here, the user is
prompted by a number of goals and assumptions in which the user is allowed to
interact with using a series of defined commands, called \textit{tactics}.
Afterwards, we show an equivalent proof using a more ``manual'' approach.

For our proof, we use the following definition of reverse:

{\small
\begin{verbatim}
rev : List a -> List a
rev [] = []
rev (x :: xs) = rev xs ++ [x]
\end{verbatim}}

To those familiar with the syntax of Haskell, this definition of {\small\tt rev}
is recursively defined for an arbitrary list ({\small\tt xs}), hence the need
for an \textit{inductive} proof. Since {\small\tt rev} is defined using
{\small\tt ++}, Idris' list append, it is helpful to first prove other properties
with respect to {\small\tt ++}. In this case, we need a proof for
\textit{the distributivity of} {\small\tt ++} \textit{over} {\small\tt rev},
which in turn requires a proof of {\small\tt ++}-\textit{identity} and the
\textit{associativity of} {\small\tt ++}.

{\small
\begin{verbatim}
rev_append_dist : (xs,ys : List a) -> 
    rev (xs ++ ys) = rev ys ++ rev xs
append_id : (xs : List a) -> xs ++ [] = xs
\end{verbatim}}

We provide these proofs in full after completing the double-reverse theorem.
Since Idris has {\small\tt appendAssociative}, we do not include a proof
of the associativity of {\small\tt ++}.

Now we are ready to start the interactive prover. To start, we must first
allow Idris to create a pattern-match statement for {\small\tt drev\_thm}.

{\small
\begin{verbatim}
drev_thm xs = ?drev_thm_rhs
\end{verbatim}
}

Here, {\small\tt ?drev\_thm\_rhs} represents an \textit{incomplete proof}.
We now have the option of completing the proof using the interactive prover.
The proof starts with no assumptions and a single goal.

{\small
\begin{verbatim}
{hole0} : (a : Type) -> (xs : List a) ->
   rev (rev xs) = xs
\end{verbatim}
}

The only valid tactic for this step is {\small\tt intros}, which creates a
new goal and moves all expressions to the left of an \(\rightarrow\) into the
proof's assumption list.

{\small
\begin{verbatim}
Assumptions: (a : Type), (xs : List a)
{hole2} : rev (rev xs) = xs
\end{verbatim}
}

From these assumptions, it is clear that {\small\tt xs} is a list. Knowing
this, the next valid tactic is {\small\tt induction xs}, allowing us to
perform induction on the list {\small\tt xs}. This command creates 2
subgoals: \textit{base} and \textit{induction}. Peforming induction on a list
reveals its \textit{constructors}, {\small\tt Nil} and {\small\tt (::)}
(i.e., empty list and \textit{cons}). By default, we must first prove the base
subgoal.

{\small
\begin{verbatim}
elim_Nil0 : rev (rev []) = []
\end{verbatim}
}

Idris can \textit{normalize} this goal using the {\small\tt compute} tactic,
resulting in {\small\tt [] = []}. Since this expression \textit{holds by reflexivity},
using the {\small\tt trivial} tactic completes base subgoal.

{\small
\begin{verbatim}
elim_::0 : (t__0 : a) -> (l__0 : List a) -> 
    (rev (rev l__0) = l__0) ->
    rev (rev (t__0 :: l__0)) = t__0 :: l__0
\end{verbatim}
}

The above expression is the start of the inductive subgoal, which gives us an
\textit{inductive hypothesis} ({\small\tt IH}). Here, we can use a specialized
{\small\tt intro} tactic to name to our inductive hypothesis and as well
as rename the variables {\small\tt t\_\_0} and {\tt l\_\_0}. Here, we
name our inductive hypothesis {\tt IH} and rename {\tt t\_\_0} and {\tt l\_\_0}
as {\tt y} and {\tt ys}, respectively.

{\small
\begin{verbatim}
Apply (intro y,ys,ih) =>
Assumptions: (a : Type), (xs : List a), (y : a), 
    (ys : List a), (IH : rev (rev ys) = ys)
{hole11} : rev (rev (y :: ys)) = y :: ys
\end{verbatim}
}

We normalize this subgoal using {\small\tt compute}.

{\small
\begin{verbatim}
{hole11} rev (rev ys ++ [y]) = y :: ys
\end{verbatim}
}

Next, we use the proof of {\small\tt rev\_append\_dist} on the lists
{\small\tt (rev ys)} and {\small\tt [y]}. To do this, we apply
{\small\tt rewrite sym (rev\_append\_dist (rev ys) [y])} to the above
subgoal. This simplifies {\small\tt rev (rev ys ++ [y])} with respect to
\textit{symmetry}.

{\small
\begin{verbatim}
{hole12} : y :: rev (rev ys) = y :: ys
\end{verbatim}
}

Now, we have a subgoal that we can rewrite using our inductive hypothesis
({\small\tt IH}), {\small\tt rev(rev ys) = ys}. We do this by using
{\small\tt rewrite sym IH}, resulting in another expression that holds by reflexivity.

{\small
\begin{verbatim}
{hole29} : y :: ys = y :: ys
\end{verbatim}
}

Applying {\small\tt trivial} finishes the proof. Using the {\small\tt qed}
tactic gives the option of saving our proof (i.e. our tactics). Below are the proofs
for {\small\tt append\_id}, {\small\tt rev\_append\_dist} and {\small\tt drev\_thm}.

{\small
\begin{verbatim}
drev_thm
    intros
    induction xs
    compute
    trivial
    intro y,ys,IH
    compute
    rewrite sym (rev_append_dist (rev ys) [y])
    rewrite sym IH
    trivial
rev_append_dist
    intros
    induction xs
    compute
    induction ys
    trivial
    intro z,zs,IH
    compute
    rewrite (append_id (rev zs ++ [z]))
    trivial
    intro z,zs,IH
    compute
    rewrite sym IH
    rewrite appendAssociative (rev ys) (rev zs) [z]
    trivial
append_id
    intros
    induction xs
    compute
    trivial
    intro y,ys,IH
    compute
    rewrite sym IH
    trivial
\end{verbatim}
}

Essentially, using the interactive prover allows us to construct inhabitants
for the types of our theorems. We may choose to do a similar proof using
only \textit{rewrite} rules. This makes our proofs more compact and gives
us the feeling of constructing them from the ground up in a more familiar
fashion. For example, the complete proof for {\small\tt drev\_thm} using
only {\small\tt rewrite} rules is given below.

{\small
\begin{verbatim}
1| drev_thm : (xs : List a) -> rev (rev xs) = xs
2| drev_thm [] = Refl
3| drev_thm (x :: xs) = 
     rewrite rev_append_dist (rev xs) [x] in
       rewrite drev_thm xs in Refl
\end{verbatim}
}

To generate this proof, we need a type to prove (viz. first line) and
pattern match expressions (viz. second and third line). Here, we use
two pattern match expressions, one for \textit{base} and \textit{inductive}.
From here, we have two incomplete proofs:

{\small
\begin{verbatim}
drev_thm : (xs : List a) -> rev (rev xs) = xs
drev_thm [] = ?drev_thm_rhs_1
drev_thm (x :: xs) = ?drev_thm_rhs_2
\end{verbatim}
}

Since our proof uses the propositional connective {\small\tt =}, we are
required to build up proofs on {\small\tt Refl}. Thus, we are finished
with a given proof only when the goal we are presented with is provable using
{\small\tt Refl} (i.e., the goal is of the form {\small\tt x = x}, for any
arbitrary expression {\small\tt x}.).

In this proof method, we are free to choose whichever case to prove first, since
in both cases the proofs are independent of each other. Here, we start with the
base case. The base case ({\small\tt ?drev\_thm\_rhs\_1}) gives us the goal:

{\small
\begin{verbatim}
`--          a : Type
-------------------------
drev_thm_rhs_1 : [] = []
\end{verbatim}
}

Since {\small\tt [] = []} is provable by {\small\tt Refl}, we write \((2)\) above.
The inductive case ({\small\tt ?drev\_thm\_rhs\_2}) requires a bit more work.

{\small
\begin{verbatim}
`--          a : Type
             x : a
            xs : List a
----------------------------------------------
drev_thm_rhs_2 : rev (rev xs ++ [x]) = x :: xs
\end{verbatim}
}

Here, the expression {\small\tt rev (rev xs ++ [x])} can be simplified using
a proof of {\small\tt rev\_append\_dist} on {\small\tt (rev xs)} and {\small\tt [x]},
resulting in the updated goal below.

{\small
\begin{verbatim}
`--          a : Type
             x : a
            xs : List a
 _rewrite_rule : rev [x] ++ rev (rev xs) = 
                   rev (rev xs ++ [x])
--------------------------------------------
drev_thm_rhs_2 : x :: rev (rev xs) = x :: xs
\end{verbatim}
}

{\small\tt \_rewrite\_rule} is the proof of {\small\tt rev\_append\_dist}
on the lists {\small\tt (rev xs)} and {\small\tt [x]}. From here, the only
valid step is to perform induction on {\small\tt xs}. In this proof method,
we perform induction and the rewriting of an inductive hypothesis with a
single {\small\tt rewrite} rule. We use {\small\tt rewrite drev\_thm xs}
giving us the assumption that {\small\tt xs = rev (rev xs)} holds,
resulting in the expression provable by {\small\tt Refl} given below.

{\small
\begin{verbatim}
`--          a : Type
             x : a
            xs : List a
 _rewrite_rule : rev [x] ++ rev (rev xs) = 
                  rev (rev xs ++ [x])
_rewrite_rule1 : xs = rev (rev xs)
------------------------------------------
drev_thm_rhs_2 : x :: xs = x :: xs
\end{verbatim}
}

We then write \((3)\) in the finished proof above for {\small\tt drev\_thm}.
For more examples using this proof method, we include the proofs of
{\small\tt rev\_append\_dist} and {\small\tt append\_id} below.
For clarity, we follow the same formatting used in the complete proof
of {\small\tt drev\_thm} (e.g. \((1)\) is the type statement,
\((2)\) is the proof of the base case, \((3)\) is the proof of the
inductive/recursive case).

{\small
\begin{verbatim}
1| append_id : (xs : List a) -> xs ++ [] = xs
2| append_id [] = Refl
3| append_id (x :: xs) = 
     rewrite append_id xs in Refl

1| rev_append_dist : (xs,ys : List a) -> 
     rev (xs ++ ys) = rev ys ++ rev xs
2| rev_append_dist [] ys = 
     rewrite append_id (rev ys) in Refl
3| rev_append_dist (x :: xs) ys = 
     rewrite rev_append_dist xs ys in 
       rewrite appendAssociative (rev ys) 
                                 (rev xs) 
                                 [x] in Refl
\end{verbatim}
}

In these proofs, the order that we apply {\small\tt rewrites} does not matter.
We can see the reason for this when we look at the final goal for
{\small\tt drev\_thm}. The expressions above the line are assumptions in
\textit{context}. Their order only matters when expressions are dependent of
each other. For example, the expression \mbox{{\small\tt (a : Type)}} must
come before {\small\tt (x : a)} and \mbox{{\small\tt (xs : List a)}} since
{\small\tt a} appears in these expressions and guarantees that {\small\tt a}
is a type. On the other hand, both {\small\tt rewrite} rules do not
\textit{introduce} any new names or expressions into context. Instead, they behave,
more or less, like proofs that allow us to construct new goals. This is
ideal since this maintains the correspondence between \textit{programs} and
\textit{proofs}. That is, in either case, the manner that we choose to (correctly)
evaluate or reduce them does not (and should not) affect the final outcome.

%%%% HoTT Route
Write a better segue.
%% From here, we are now familiar with handling inductive types, one of which is type of
%% \textit{equality}. Briefly, equality is built up with a single constructor, \textit{refl}.
%% Refl, or \textit{reflexivity}, states that an expression is only equal to itself, similar to
%% Liebniz's Law. We can see this in Idris' interactive prover when applying the tactic
%% {\small\tt trivial}, since this tactic attempts to reduce a goal with respect to reflexivity.
%% When \(\lambda C\) is extended with induction and the notion of equality, it is possible
%% to extend it even further to a system that simulates \textit{Homotopy Type Theory}.

\subsection*{Homotopy Type Theory}
Equivalence is a proof of equality, or equivalence is equivalent to equality. Equality is
path equality (homotopical). Two types are equivalent, when they are equivalent with
respect to an existence between their elements.

%%%% Co-induction Route
%% \par\hspace{2ex}  
%% In a co-inductive system, we can only prove properties with respect to \textit{productivity} and \textit{bisimilarity}.
%% In other words, we can only prove properties on the behavior of infinite data by appealing to the manner in which they are built up.
%% \subsubsection*{Co-inductive Constructions}
%% Extending \(\lambda C\) with co-inductive types allows an encoding of infinite data structures (e.g. lazy lists or \textit{streams}),
%% resulting in the system of \textit{The Calculus of Co-inductive Constructions} \cite{unknown}.
%% \par\hspace{2ex}  
%% We, once again, use Idris to model a simple co-inductive proof. Since Idris is built on Haskell, it already has access to
%% streams \cite{unknown}. Unfortunately, Idris' interactive prover does not yet fully support co-inductive proofs,
%% however, we can still prove certain properties of infinite data by appealing to induction.

\section* {Final Remarks}
\begin{itemize}
\item Talk about \(\lambda C\) and \textit{why} it is a good foundation for
  learning and modelling Type Theory
\item Specifically, it can be taught in bits and pieces from its components,
  which comprise the majority of the ideas of type theory
\item Also, since Type theory is meant to be \textit{computable}, it should be
  modelled by something that can compute. \(\lambda C\) does this since it is a
  system built on computation (i.e., a calculus). Effectively, this gives rise to
  something that can be simulated by a computer program (i.e., all features of type
  theory can be prescribed an algorith, since it can be modelled entirely by \(\lambda C\)).
\item Briefly talk about HoTT and its relationship with Type theory and how it
  is essentially the new foundation for mathematics (maybe talk about how it supports
  both classical and constructive logic?)
\end{itemize}
%% The Calculus of Constructions is great system to start out with in learning Type Theory. It, however, should not be the stopping point.
%% In fact, there are already two variants of the calculus that are more powerful and expressive: the Calculus of Inductive Constructions
%% and the Calculus of Co-inductive Constructions. The former is the system that the proof assistant Coq is built on and is a requirement
%% to add \textit{recursion} and an encoding of \textit{peano} Natural numbers.
%% The later is a system that permits \textit{infinite data types} (i.e. lazy lists or \textit{streams}) and
%% has the fascinating ability to prove properties about them. Functional programming would not be
%% as expressive and powerful as it is today without these systems.
``Type Theory and Formal Proof'' offers insight on a variant of \(\lambda C\) called
\(\lambda D\). \(\lambda D\) is a system that has similar power to the Calculus of
Inductive Constructions by extending \(\lambda C\) with \textit{definitions}. In effect,
\(\lambda D\) retains the relative simplicity of the base \(\lambda C\) system,
while also simulating the power of the more complicated \(\lambda C\) variants.

Write a better final paragraph. Don't include opinion.
%% In this paper, we simply offer our own perspective of an ``introduction'' to Type Theory with hopes of inspiring
%% individuals to wade into the field. If they were to do so, we hope that they would find as much fascination and
%% intrigue as many of the forerunners of the field have found in it.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{1}
  %% HP Barendregt 1984
  %% HP Barendregt 1992
  %% Church 1940
  %% Thierry Coquand 1988
  %% Girard 1972
  %% Martin-Lof 1980
  %% Nederpelt and Geuvers 2014
  %% Sanchis

\bibitem{bar1} Henk Barendregt. {\em The Lambda Calculus: Its Syntax and Semantics}
  1984: Elsevier Science.

\bibitem{bar2} Henk Barendregt. {\em Lambda Calculi with types}
  1992: Handbook of Logic in Computer Science, pp. 117-309, Oxford University Press.

\bibitem{bert} Yves Bertot and Pierre Cast\'eran. {\em Interactive Theorem Proving and Program
  Development: Coq\'Art: the Calculus of Inductive Constructions} 2004: Springer.

\bibitem{church} Alonzo Church. {\em A formulation of the simple theory of types}
  1940: Journal of Symbolic Logic, 5, pp. 56-68.

\bibitem{coq1} Thierry Coquand and G\'erard Huet. {\em The Calculus of Constructions}
  1988: Information and Computation, 76, pp. 95-120.

\bibitem{gir} Jean-Yves Girard. {\em Interpr\'etation fonctionnelle et \'elimination des coupures de l’arithm\'etique d’ordre sup\'erieur}
  1972: PhD thesis, Universit\'e Paris VII.

\bibitem{gir2} Jean-Yves Girard, Yves Lafont, and Paul Taylor. {\em Proofs and Types}
  1989: Cambridge University Press, pp. 82-86.

\bibitem{how} William Howard. {\em The formulae-as-types notion of construction}
  1980: Seldin and Hindley, pp. 479-490.

\bibitem{lof} Per Martin-L\"of. {\em Intuitionistic Type Theory}
  1980: Bibliopolis

\bibitem{rng} Rob Nederpelt and Herman Geuvers. {\em Type Theory and Formal Proof}
  2014: Cambridge University Press.

\bibitem{san} Luis Sanchis. {\em Functionals defined by Recursion}
  1967: Notre Dame Journal of Formal Logic VIII, pp. 161-174.

\end{thebibliography}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2016-07-03 paper.tex
yo

@@
 
-\begin{abstract}
-  Who knows.
-\end{abstract}
+%% \begin{abstract}
+%% \end{abstract}
 
 \section{Introduction}
-Introduce the contents of the paper, notation, talk about Agda.
+We present an introduction to dependent types and fundamental type theory.
+At its core, type theory seeks to serve as a possible foundation of mathematics,
+which itself is based on a different logical foundation. This proof-relevant
+form of mathematics is the basis in which types can be used to model logical
+propositions and is the core idea of type theory. As such, not only does type
+theory have a viable application in mathematics but also in computer science,
+providing a richer world of types and expressibility.
+
+The study of type theory goes hand in hand with the study of dependent types,
+which is the reason why first take the time to do so in Section 2. After we
+have derived a proper foundation for type theory, we then continue with the
+foundational concepts of Martin-L\"of type theory. We recommend using a proof
+assistant while working and studying with the examples we provide. For such
+examples, we use Agda notation.
 
@@
 \end{enumerate}
-These dependecies, however, should not be confused with an actual dependent type.
-In fact, only (2) and (4) correspond to expressions that are typed (necessarily)
-with a dependent type (we explain later why (1) and (3) are not typed with
-dependent types).
+These dependecies, however, should not be confused with an actual dependent
+type. In fact, only (2) and (4) correspond to expressions that are typed
+(necessarily) with a dependent type (we explain later why (1) and (3) are
+not typed with dependent types).
 
@@
 \end{center}
-The terms of the STLC are simply the terms of the \textit{untyped \(\lambda\)-calculus},
+The terms of the STLC are the terms of the \textit{untyped \(\lambda\)-calculus},
 with the exception of terms that use \textit{self-application} (i.e., Y
 combinator, Omega, etc.). This has the benefit of removing all \textit{infinite}
-calculations from the system but also removing all recursive computations. Thus,
-the STLC is not \textit{Turing complete} and is thus the appropriate foundation
-for deriving a type theoretic system. It must first, however, be extended with
-more expressive types, as the only form of dependency present in the STLC is (1),
-terms depending on terms (i.e. {\em term level} \(\lambda\)-abstraction).
+calculations from the system but also removing all recursive computations \cite{}.
+Thus, the STLC is not \textit{Turing complete} and is thus the appropriate
+foundation for deriving a type theoretic system. This is because the STLC lends
+itself quite seamlessly as component of a larger, more expressive system. This
+flexibility is what we exploit to derive \(\lambda C\). To do this, we first
+extend the STLC with more expressive types through simple extentions on its
+grammar for simple types.
+
+%% as the only form of dependency
+%% present in the STLC is (1), terms depending on terms (i.e. {\em term level}
+%% \(\lambda\)-abstraction).
 
@@
 (viz. definition of simple type). For example, the \textit{constant function}
-over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC looks like:
+over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC is defined:
 \begin{center}
@@
 
-Another name for \(\lambda2\) is System F. For our purposes, we treat
+Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
 \(\lambda2\) as simply an extension of the STLC. This system is adds
@@
 type (e.g. {\tt (x : A) → ...}). Types that appear in this fashion are
-called \(\Pi\)-types. This leads us to a new grammar for types ({\tt t}):
+called \(\Pi\)-types. Taking the grammar for simple types, we can add \(\Pi\)-types,
+resulting in the grammar for types ({\tt t}) in \(\lambda2\):
 
@@
 \(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
-over other types (e.g. {\tt t} in {\tt ∀ (x : Set) → t}). In a later system, we
+over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). In a later system, we
 allow \(\Pi\)-types to be parameterized over terms (e.g. {\tt ∀ (x : t) → t}).
@@
 \begin{minipage}{0.4\textwidth}
-STLC \(\rightarrow\) \(\lambda2\) \(\rightarrow\) \(\lambda\underline{\omega}\) \(\rightarrow\) \(\lambda\omega\)
+STLC \(\rightarrow\) \(\lambda2\) \(\rightarrow\) (+\(\lambda\underline{\omega}\)) \(\rightarrow\) \(\lambda\omega\)
 \end{minipage}
@@
 We are already familiar with the first two levels, terms and types, which, from
-the STLC, are still related in terms of inhabitance. To meet the constraints of
-the STLC when introducing type constructors to our system, we must also provide
-the type of types. For our purposes, we assume an infinite universe of types, as
-is common practice in deriving type theoretic systems to avoid certain paradoxes.
-For example, let's first assume that natural numbers are primitive within our
-system. We then have the following for the natural number 42:
+the STLC, are related in terms of inhabitance. To introduce the third level and
+onward, we must also provide the type of types. For our purposes, we assume an
+infinite universe of types, as is common practice in deriving type theoretic
+systems to avoid certain paradoxes \cite{}. For example, let's first assume that
+natural numbers are primitive within our system. We then have the following for
+the natural number 42:
 
@@
 
-Another name for {\tt Set} is a {\em kind}. For our puroses, we use the
-more familiar set-theoretic equivalent reading. Reading the above example is rather
-straightforward: 42 is of type \(\mathbb{N}\) which is of type {\tt Set}
-which of type {\tt Set₁} and so on. There is, however, an important distinction
-for the first two levels: 42 is NOT of type {\tt Set}, while \(\mathbb{N}\) is
-both of type {\tt Set} and {\tt Set₁}. Terms are therefore treated uniquely,
-since they only exist as elements of types. While important, this topic is a bit
-out of the scope of an introductory paper, however, we briefly discuss this
-in a later section.
+We read the above example as: 42 is of type \(\mathbb{N}\) which is of type
+{\tt Set} which of type {\tt Set₁} and so on. There is, however, an important
+distinction for the first two levels: 42 is NOT of type {\tt Set}, while
+\(\mathbb{N}\) is both of type {\tt Set}, {\tt Set₁} and onwards. Terms are
+therefore treated uniquely, since they only exist as elements of types.
 
@@
 {\tt t → t}, where either {\tt t} can have subexpressions containing {\tt Set}.
-Let's take a familiar definition of the list data structure:
+Let's take the familiar definition of the list data structure:
 
@@
 {\tt A}. Another way to think about this statement is to think of {\tt List} as
-a form of polymorphic function of type {\tt Set → Set}, as is similar to other
-{\em parameterized types}, that can be applied at type level (thus the name
-{\em type constructor}). This is to insure that we cannot have create type
-expressions that fail to instantiate {\tt List} (i.e., {\tt List → A}).
+a form of polymorphic function of type {\tt Set → Set}, as is similar with other
+{\em parameterized types}. Unlike like {\tt const} or {\tt polyConst}, this
+function can only be applied at type level (thus the name {\em type constructor}).
+This is to insure that we cannot write type expressions that fail to instantiate
+{\tt List} (e.g. {\tt List → A} v. {\tt List A → A}).
 
-From this we can see that the simple addition of {\tt Set} into our grammar of
+From this we can see that the addition of {\tt Set} into our grammar of
 types has great effect on the expressivity of our system. There is only one
@@
 system of \(\lambda P\), allows us to construct {\em type families} which are
-meant to directly correspond to propositions in higher-order logic.
+types themselves and are meant to directly correspond to propositions in
+higher-order logic.
 
@@
 a natural number. Another way to say this is that {\tt isEven} is a predicate
-over natural numbers. When writing propositions, we write them as the equivalent
+over natural numbers. When writing propositions, we write them as their equivalent
 type definition (similar to {\tt List}). Consequently, the proposition
 {\tt isEven} is an example of a dependent type, since its {\em meaning} depends
-on the natural number passed to it.
+on the natural number passed to it (types depending on terms).
 
@@
 
-This is the last form of dependency of the four we mentioned earlier. Here,
-we've defined two ways in which {\tt n isEven} for all natural numbers
-({\tt n}) can be proven ``true'', which is done via its two type constructors
-{\tt ZEven} and {\tt NEven}, corresponding to the two constructors of
-\(\mathbb{N}\). {\tt ZEven} corresponds to the proof that 0 is an even number.
-{\tt NEven} then defines the property that if a given natural number ({\tt n}) is
-even (viz. {\tt n isEven}), then {\tt n + 2} is even (viz. {\tt (Suc (Suc n))
-isEven}). To clarify how this type can be used, let's prove that 4 is even. We
-start by providing the appropriate type definition and also providing a name for
-our proof (viz. {\tt 4isEven}):
+Here, we've defined two ways in which {\tt n isEven} for all natural numbers
+({\tt n}) can be proven. Returning to the idea of proof-relevant mathematics,
+we do this by use of its two type constructors {\tt ZEven} and {\tt NEven},
+corresponding to the two constructors of \(\mathbb{N}\). {\tt ZEven} is the
+proof that 0 is an even number. {\tt NEven} then defines the property that if
+a given natural number ({\tt n}) is even (viz. {\tt n isEven}), then {\tt n + 2}
+is even (viz. {\tt (Suc (Suc n)) isEven}). To clarify how this type can be used,
+let's prove that 4 is even. We start by providing the appropriate type definition
+and also providing a name for our proof (viz. {\tt 4isEven}):
 \begin{center}
@@
 \end{center}
-We must then show that 4 is indeed even using a combination of {\tt ZEven} and
-{\tt NEven}. From the definition of {\tt isEven}, we know that to prove 4 is
-even, we must know that 2 is even.  Consequently, to prove 2 is even, we must
-know that 0 is even. Fortunately, we have a proof that 0 is even: {\tt ZEven}.
-Thus, the proof that 2 is even is {\tt NEven 0 ZEven}, and finally the proof
-that 4 is even is:
+From the definition of {\tt isEven}, we know that to prove 4 is even, we must
+know that 2 is even. Consequently, to prove 2 is even, we must know that 0 is
+even. Fortunately, we have a proof that 0 is even: {\tt ZEven}. Thus, the proof
+that 2 is even is {\tt NEven 0 ZEven}, and finally the proof that 4 is even is:
 \begin{center}
@@
 
-This method of proof is adopted from {\em Intuitionistic Logic}. In this form of
-logic, ``truth'' corresponds to the idea of {\em type inhabitability}. This
-makes sense since we use define types to correspond to logical propositions
-(e.g. {\tt isEven}). ``Falsity'' then naturally corresponds to un-inhabitability
+This method of proof is adopted from {\em Intuitionistic Logic} \cite{}. In
+this form of logic, ``truth'' corresponds to the idea of {\em type inhabitability}.
+This makes sense since we use types to correspond to logical propositions
+(e.g. {\tt isEven}). ``Falsity'' then naturally corresponds to the un-inhabitability
 of a given type. This can be seen if we try to prove that 5 is even using
-{\tt isEven}. The proof seems to get ``stuck.''
+{\tt isEven}. The proof seems to get ``stuck'' at trying to prove that 1 is even.
 
@@
 the relationship between each of the systems we have previously mentioned, we
-include a diagram of the {\em Lambda Cube}:
+include a diagram of the {\em Lambda Cube} \cite{}:
 \begin{center}
@@
 \end{center}
-Reading the cube requires us to start on the edge marked \(\lambda\rightarrow\),
+Reading the cube requires us to start on the edge marked \(\lambda\!\!\rightarrow\),
 which corresponds to the STLC, the base system. From here, we can either move

### 2016-07-03 paper.tex
yo

@@
 \DeclareUnicodeCharacter{955}{\ensuremath{\lambda}}
+\DeclareUnicodeCharacter{2261}{\ensuremath{\equiv}}
 
@@
 In this section, we introduce the foundational concepts of Martin-L\"of Type
-Theory. We discuss recursion, induction, and as well several primitive type
-encodings (e.g. from propositional logic). We recommend the reader to have a
-running proof assistant while reading this section to help fortify (better word)
-several concepts.
+Theory \cite{}. We discuss recursion, induction, and as well several primitive type
+encodings (e.g. from propositional logic).
 
@@
 because, in general, recursion and induction principles are defined under certain
-known intuitions on the given type that they are meant to act on and thus also
-follow a certain pattern.
+known intuitions (we can think of them as axioms) on the given type that they are
+meant to act on. They also follow a certain pattern:
 \begin{center}
@@
 \begin{small}
-
 The recursion and induction principles for some type {\tt T}.
@@
 %% Bottom, Top, Disjunction, Conjunction, Sigma
+We start with \(\perp\), the empty type. We include the definition of \(\perp\),
+{\tt rec⊥} and {\tt ind⊥}:
 
@@
 \subsection*{Booleans}
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+data 𝔹 : Set where
+  True  : 𝔹
+  False : 𝔹
+
+rec𝔹 : (C : Set) → C → C → 𝔹 → C
+rec𝔹 C t f True = t
+rec𝔹 C t f False = f
+
+ind𝔹 : (C : 𝔹 → Set) → C True → C False → (x : 𝔹) → C x
+ind𝔹 C t f True = t
+ind𝔹 C t f False = f
+\end{code}
+\end{minipage}
+\end{center}
+
 \subsection*{Natural Numbers}
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+recℕ : (C : Set) → C → (ℕ → C → C) → ℕ → C
+recℕ C c _ 0 = c
+recℕ C c f (suc n) = f n (recℕ C c f n)
+
+indℕ : (C : ℕ → Set) → C 0 →
+       (∀ (x : ℕ) → C x → C (suc x)) →
+       (x : ℕ) → C x
+indℕ C c _ 0 = c
+indℕ C c f (suc n) = f n (indℕ C c f n)
+\end{code}
+\end{minipage}
+\end{center}
+
 \subsection*{List/Orderings}
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+data List (A : Set) : Set where
+  Nil  : List A
+  _::_ : A → List A → List A
+
+recList : {A : Set} → (C : Set) → C →
+          (A → List A → C → C) → List A → C
+recList C c f Nil = c
+recList C c f (x :: xs) = f x xs (recList C c f xs)
+
+indList : {A : Set} → (C : List A → Set) → C Nil →
+          ((x : A) → (xs : List A) → C xs → C (x :: xs)) →
+          (xs : List A) → C xs
+indList C c f Nil = c
+indList C c f (x :: xs) = f x xs (indList C c f xs)
+\end{code}
+\end{minipage}
+\end{center}
+
 \subsection*{Identity Types}
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+infix  4  _≡_     -- propositional equality
+
+data _≡_ {ℓ} {A : Set ℓ} : (a b : A) → Set ℓ where
+  refl : (a : A) → (a ≡ a)
+
+ind≡ : {A : Set} (C : {x y : A} → x ≡ y → Set) →
+       (∀ (x : A) → C (refl x)) →
+       {x y : A} (p : x ≡ y) → C p
+ind≡ C c (refl x) = c x
+
+rec≡ : {A : Set} {x y : A} (C : A → Set) → (p : x ≡ y) → C x → C y
+rec≡ {A} {x} {y} C = 
+  ind≡ (λ {x} {y} _ → C x → C y)
+       (λ x → λ z → z)
+       {x} {y} 
+\end{code}
+\end{minipage}
+\end{center}
 

### 2016-07-04 paper.tex
yo

@@
 \section{Example Types and Sample Proofs}
-Describe the section.
+In this section, we present the type of Booleans, Natural numbers, Lists,
+and the identity type. To further understand these types, we also include
+a brief description of the different forms of equality, namely definitional,
+judgmental and propositional. In type theory, the specific differences
+between the three are generally misunderstood but are important in understanding
+basic type theoretic proof methods, which we also go into some detail in
+this section with several examples.
+
+\subsection*{Equality}
+%% talk about definitional equality, judgmental equality, prop equality
 
@@
 \end{center}
+%% proof: 1 + 1 = 2
 
@@
 \subsection*{Identity Types}
+%% say something that this is the HoTT version of identity
+%% possibly also point out that the discrepancy between the two
 \begin{center}
@@
 \begin{code}
-infix  4  _≡_     -- propositional equality
-
+infix  4  _≡_
 data _≡_ {ℓ} {A : Set ℓ} : (a b : A) → Set ℓ where
@@
 
-ind≡ : {A : Set} (C : {x y : A} → x ≡ y → Set) →
+-- "PROPOSITIONAL EQUALITY"
+-- For a proposition on an arbitrary identity type,
+-- if we can show that it holds for (refl x), then
+-- C holds for all proofs, p, such that x ≡ y.
+J : {A : Set} (C : {x y : A} → x ≡ y → Set) →
     (∀ (x : A) → C (refl x)) →
     {x y : A} (p : x ≡ y) → C p
-ind≡ C c (refl x) = c x
+J C c (refl x) = c x
 
-rec≡ : {A : Set} {x y : A} (C : A → Set) → (p : x ≡ y) → C x → C y
+-- "STRICT EQUALITY"
+-- For a proposition on a reflexive equality (x ≡ x),
+-- if we can show that it holds for (refl x), then
+-- C holds for all other reflexive equalities p : x ≡ x.
+K : {A : Set} (C : {x : A} → x ≡ x → Set) →
+    (∀ (x : A) → C (refl x)) →
+    {x : A} (p : x ≡ x) → C p
+K C c (refl x) = c x
+
+rec≡ : {A : Set} {x y : A} (C : A → Set) →
+       (p : x ≡ y) → C x → C y
 rec≡ {A} {x} {y} C = 
-  ind≡ (λ {x} {y} _ → C x → C y)
+  J (λ {x} {y} _ → C x → C y)
     (λ x → λ z → z)

### 2016-07-05 paper.tex
moe wok

@@
 Theory \cite{}. We discuss recursion, induction, and as well several primitive type
-encodings (e.g. from propositional logic).
+encodings (from propositional logic).
 
@@
 Recursion and induction are one of the powerful tools available in a
-fully-dependently typed system. Using recursion and induction principles to
-define functions and proofs, we are guaranteed termination and totality. This is
-because, in general, recursion and induction principles are defined under certain
-known intuitions (we can think of them as axioms) on the given type that they are
-meant to act on. They also follow a certain pattern:
+fully-dependently typed system. Recursion and induction principles can be
+thought of as specialized functions to behave on elements of a given type.
+If we use recursion and induction principles to define functions and proofs,
+we are guaranteed both termination and totality. This is because, in general,
+recursion and induction principles are defined under certain known intuitions
+(which is why they are sometimes considered axioms) on the given type that
+they are meant to act on.
+
+These principles follow a certain pattern:
 \begin{center}
@@
 The first argument given to these principles is sometimes called the {\em motive}.
-This simply means that {\tt C} is the type we are going to inhabit. For recursion
-principles, this means the given output of the function we are defining, and for
-induction principles, this can be read as the proposition we are trying to prove.
-The terms that follow (viz. {\tt C₀}) correspond to the constructors of the given
-type, which is how these principles guarantee totality. Finally, both principles
-take a term of the type that are acting on (viz. {\tt T} and {\tt (t : T)}). The
-only difference is that the induction principle introduces a type family, as
-opposed to non-parameterized type in the recursion principle. We discuss this in
-Section 4.
+This simply means that {\tt C} is the type from which we will return a element
+from. For recursion principles, this means the given output of the function we
+are defining, and for induction principles, this can be read as the proposition
+we are trying to prove. The terms that follow (viz. {\tt C₀}) correspond to the
+constructor(s) of the given type, which is how these principles guarantee
+totality. Finally, both principles take a term of the type that are acting on
+(viz. {\tt T} and {\tt (t : T)}). The only difference is that the induction
+principle introduces a type family, as opposed to a non-parameterized type in the
+recursion principle. We discuss this in Section 4.
 
@@
 %% Bottom, Top, Disjunction, Conjunction, Sigma
-We start with \(\perp\), the empty type. We include the definition of \(\perp\),
-{\tt rec⊥} and {\tt ind⊥}:
+We start with \(\perp\), the empty type. Since there no ways to construct
+an element of the empty type, the encoding is rather simple:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+data ⊥ : Set where
+
+rec⊥ : (C : Set) → ⊥ → C
+rec⊥ C ()
+
+ind⊥ : (C : ⊥ → Set) → (z : ⊥) → C z
+ind⊥ C ()
+
+prop : {A : Set} → ⊥ → A
+prop {A} perp = rec⊥ A perp
+
+prop' : {A : ⊥ → Set} → (x : ⊥) → A x
+prop' {A} perp = ind⊥ A perp
+\end{code}
+\end{minipage}
+\end{center}
+The behavior of the empty type is meant to model {\it ex-falso} (i.e., from
+absurdity comes anything we like). In general, we use the empty type when
+writing proofs containing negated expressions.
+
+To clarify our discussion about the empty type, we include a proof of:
+$$\neg\neg\neg A\rightarrow\neg A$$
+Here, we define negation, \(\neg\), to be a function, such that:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+¬ : Set → Set
+¬ A = A → ⊥
+
+prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
+prop₀ ¬¬¬a a = ?
+\end{code}
+\end{minipage}
+\end{center}
+Now, we solve for {\tt ?}. Looking at the type expression of {\tt prop₀},
+we know we have an expression of type \(\neg\neg\neg A\) and also an
+expression of type \(A\). To clarify, we can rewrite the type expression
+after desugaring \(\neg A\):
+$$\neg\neg\neg A\rightarrow A\rightarrow\perp$$
+Now, to construct an element of type \(\neg A\) from an element of type
+\(\neg\neg\neg A\), we can use the definition of \(\neg\):
+$$(\neg\neg A\rightarrow\perp)\rightarrow A\rightarrow\perp$$
+$$((\neg A\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
+$$(((A\rightarrow\perp)\rightarrow\perp)\rightarrow\perp)\rightarrow A\rightarrow\perp$$
+Thus, we know we must apply our expression \(\neg\neg\neg A\) to a function
+of type \(\neg\neg A\) and so on, such that:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+prop₀ : {A : Set} → ¬ (¬ (¬ A)) → ¬ A
+prop₀ ¬¬¬a a = ¬¬¬a (λ ¬a → ¬a a)
+\end{code}
+\end{minipage}
+\end{center}
+Or equivalently:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+prop₀ ¬¬¬a a = rec⊥ ⊥ (¬¬¬a (λ ¬a → ¬a a))
+\end{code}
+\end{minipage}
+\end{center}
+
+Next, we introduce the type \(\top\), which contains exactly one element,
+{\tt *}:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+data ⊤ : Set where
+  * : ⊤
+
+rec⊤ : (C : Set) → C → ⊤ → C
+rec⊤ C c * = c
+
+ind⊤ : (C : ⊤ → Set) → C * → (x : ⊤) → C x
+ind⊤ C c * = c
+\end{code}
+\end{minipage}
+\end{center}
+Similar to \(\perp\), \(\top\) is simple and is rather uninteresting by itself.
+There are, however, some trivial yet interesting proofs that involve the two.
+The first of these we also introduce a special kind of proof: {\it uniqueness
+of identity} (UIP) \cite{}. In this proof, we set out to determine and classify
+the identity of every element within a given type. For example, for \(\top\), we
+should be able to prove:
+$$\forall (x : \top) \rightarrow x ≡ *$$
+That is, for every element of type \(\top\), that element is equivalent to {\tt *}.
+This second expression, {\it x ≡} {\tt *}, is an example of an {\it identity type},
+which we describe later on in this section. For now, we simply say that the
+sole constructor for the identity type is {\it refl x}, for all x.
+
+Now, to prove the above proposition, we can use the induction principle for
+\(\top\), since the proposition we are trying to prove concerns all elements
+of \(\top\):
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+uip⊤ : (x : ⊤) → x ≡ *
+uip⊤ x = ind⊤ (λ top → top ≡ *) (refl *) x
+--uip⊤ * = refl *
+\end{code}
+\end{minipage}
+\end{center}
+Here, we have taken our motive to be the expression {\tt (λ top → top ≡ *)},
+which essentially repeats the information already given by our type expression.
+The second element, {\tt refl *}, is a proof that our motive holds in the event
+{\tt top} is {\tt *}, which is of type {\tt * ≡ *}. The last element {\tt x} is
+the expression initially passed to our proof of type \(\top\). Equivalently, we
+could have also written our proof as the third line above. This form of proof
+uses pattern matching and essentially asserts that the only element in \(\top\)
+is {\tt *}, which greatly simplifies our proof. Both methods of proof are
+correct, but, in general, using pattern matching is only safe when the values of
+proof expressions are not dependent one another (i.e., their values don't change
+depending on which expression we apply pattern matching to).
+
+Before we can do more proofs, we first introduce the disjunction type, +: 
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+data _+_ (A B : Set) : Set where
+  inl : A → A + B
+  inr : B → A + B
+
+rec+ : {A B : Set} (C : Set) → (A → C) → (B → C) →
+       A + B → C
+rec+ C f _ (inl a) = f a
+rec+ C _ f (inr b) = f b
+
+ind+ : {A B : Set} (C : A + B → Set) →
+       (∀ (x : A) → C (inl x)) →
+       (∀ (x : B) → C (inr x)) →
+       (x : A + B) → C x
+ind+ C f _ (inl a) = f a
+ind+ C _ f (inr b) = f b
+\end{code}
+\end{minipage}
+\end{center}
+It might come as surprise as to why the symbol for addition is generally used
+for the disjunction type. This is because another name for + is a {\it union type}.
+This type is parameterized over two other types, {\tt A} and {\tt B}, and
+creates a union between them in that every expression of type {\tt A + B} is
+either {\tt inl a}, where {\tt a : A}, or {\tt inr b}, where {\tt b : B}.
 
@@
 \begin{code}
+data ℕ : Set where
+  0 : ℕ
+  suc : ℕ → ℕ
+
 recℕ : (C : Set) → C → (ℕ → C → C) → ℕ → C
@@
 indList : {A : Set} → (C : List A → Set) → C Nil →
-          ((x : A) → (xs : List A) → C xs → C (x :: xs)) →
+          (∀ (x : A) → (xs : List A) → C xs → C (x :: xs)) →
           (xs : List A) → C xs
@@
 -- "PROPOSITIONAL EQUALITY"
--- For a proposition on an arbitrary identity type,
--- if we can show that it holds for (refl x), then
--- C holds for all proofs, p, such that x ≡ y.
+-- For a proposition, C, on a proof of x ≡ y for
+-- any x and y, if we can show that it holds for
+-- (refl x), then C holds for all proofs, p : x ≡ y.
 J : {A : Set} (C : {x y : A} → x ≡ y → Set) →
@@
 -- "STRICT EQUALITY"
--- For a proposition on a reflexive equality (x ≡ x),
--- if we can show that it holds for (refl x), then
+-- For a proposition, C, on a reflexive equality (x ≡ x),
+-- for any x, if we can show that it holds for (refl x), then
 -- C holds for all other reflexive equalities p : x ≡ x.

### 2016-07-14 paper.tex
first commit from ubuntu

@@
 \title{Dependent Types and Martin-L\"of Type Theory: An Introduction}
+%% \title{Dependent Types and Type Theory: An Introduction}
 \author{Carl Factora}
@@
 \section{Introduction}
-We present an introduction to dependent types and fundamental type theory.
+We present an introduction to dependent types and (Martin-L\"of) type theory.
 At its core, type theory seeks to serve as a possible foundation of mathematics,
@@
 theory have a viable application in mathematics but also in computer science,
-providing a richer world of types and expressibility.
+providing for both a richer world of types and expressibility.
 
 The study of type theory goes hand in hand with the study of dependent types,
-which is the reason why first take the time to do so in Section 2. After we
-have derived a proper foundation for type theory, we then continue with the
-foundational concepts of Martin-L\"of type theory. We recommend using a proof
-assistant while working and studying with the examples we provide. For such
-examples, we use Agda notation.
+which is  why we first take the time to do familiarize ourselves with them in
+Section 2. We then continue with the foundational concepts of type theory of
+Per Martin-L\"of \cite{}. In this section, while reading through the examples
+we have provided, we recommend using a proof assistant, such as Agda. In our
+examples, we refer to Agda notation.
 
@@
 it helps to think of the ways in which types and terms can \textit{depend} on
-each other. Altogether, there are four different kinds of dependencies:
+one another. Altogether, this results in four dependencies:
 
@@
 introduce the (fully-dependent) system of the \textit{Calculus of Constructions}
-(\(\lambda C\)). Here, the word \textit{fully} signifies that \(\lambda C\)
+(\(\lambda C\)) \cite{}. Here, the word \textit{fully} signifies that \(\lambda C\)
 features all four forms of dependency mentioned above.
@@
 its namesake. A simple type is either a {\em primitive type}, which is the set
-of all types ``pre-defined'' within the system, or a {\em function type} between
-other simple types. Thus, simple types have the following grammar:
+of all types ``pre-defined'' (e.g. lists, numbers, booleans), or a
+{\em function type} between other simple types. Thus, simple types have the
+following grammar:
 \begin{center}
@@
 with the exception of terms that use \textit{self-application} (i.e., Y
-combinator, Omega, etc.). This has the benefit of removing all \textit{infinite}
-calculations from the system but also removing all recursive computations \cite{}.
-Thus, the STLC is not \textit{Turing complete} and is thus the appropriate
-foundation for deriving a type theoretic system. This is because the STLC lends
-itself quite seamlessly as component of a larger, more expressive system. This
-flexibility is what we exploit to derive \(\lambda C\). To do this, we first
-extend the STLC with more expressive types through simple extentions on its
-grammar for simple types.
+combinator, Omega, etc.). An example term, annotated with its type, is generally
+written as follows:
+\begin{center}
+\begin{minipage}{0.5\textwidth}
+\begin{code}
+example : ℕ → ℕ
+example = λ (z : ℕ) → z
+\end{code}
+\end{minipage}
+\end{center}
+Or, equivalently,
+\begin{center}
+\begin{minipage}{0.5\textwidth}
+\begin{code}
+example = λ z → z
+\end{code}
+\end{minipage}
+\end{center}
+whenever the type of the argument (viz. {\tt z}) is easily inferable.
+
+This leaves the STLC rather small and limited but has the benefit of removing
+all \textit{infinite} calculations from the system \cite{}. Thus, the STLC is
+not \textit{Turing complete} and is thus the appropriate foundation for
+deriving a type theoretic system. This is because the STLC lends itself quite
+seamlessly as component of a larger, more expressive system. This flexibility
+is what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
+with more expressive types through simple extentions on its grammar for simple
+types.
 

### 2016-07-25 paper.tex
more

@@
 At its core, type theory seeks to serve as a possible foundation of mathematics,
-which itself is based on a different logical foundation. This proof-relevant
+one that is based on a different logical foundation. This proof-relevant
 form of mathematics is the basis in which types can be used to model logical
 propositions and is the core idea of type theory. As such, not only does type
-theory have a viable application in mathematics but also in computer science,
-providing for both a richer world of types and expressibility.
+theory have a viable application in mathematics but also a growing interest
+in the field of computer science. In both fields, type theory provides a richer
+world of types and expressibiilty.
 
 The study of type theory goes hand in hand with the study of dependent types,
-which is  why we first take the time to do familiarize ourselves with them in
-Section 2. We then continue with the foundational concepts of type theory of
-Per Martin-L\"of \cite{}. In this section, while reading through the examples
-we have provided, we recommend using a proof assistant, such as Agda. In our
-examples, we refer to Agda notation.
+which is why we first take the time to do familiarize ourselves with them
+(Section 2). We then continue with the foundational concepts of type theory of
+Per Martin-L\"of \cite{}. For this, we recommend using a proof assistant, such
+as Agda. In our examples, we refer to Agda syntax rules.
 
@@
 \end{enumerate}
-These dependecies, however, should not be confused with an actual dependent
-type. In fact, only (2) and (4) correspond to expressions that are typed
-(necessarily) with a dependent type (we explain later why (1) and (3) are
-not typed with dependent types).
+These dependecies, however, should not be confused with dependent types.
+In fact, only (2) and (4) correspond to expressions that are typed (necessarily)
+with a dependent type (we explain later why (1) and (3) are not typed with
+dependent types).
 
@@
 is drawn in constrast with the term {\em simple type}, where the STLC derives
-its namesake. A simple type is either a {\em primitive type}, which is the set
-of all types ``pre-defined'' (e.g. lists, numbers, booleans), or a
-{\em function type} between other simple types. Thus, simple types have the
+its namesake. A simple type ({\tt t}) is either a {\em primitive type}, {\tt Prim},
+which is the set of all types ``pre-defined'' (e.g. lists, numbers, booleans),
+or a {\em function type} between other simple types. Thus, simple types have the
 following grammar:
@@
 \begin{code}
-t || t → t
+Prim || t → t
 \end{code}
@@
 example : ℕ → ℕ
-example = λ (z : ℕ) → z
+example = λ (n : ℕ) → n
 \end{code}
@@
 \begin{code}
-example = λ z → z
+example = λ n → n
 \end{code}
@@
 \end{center}
-whenever the type of the argument (viz. {\tt z}) is easily inferable.
+whenever the type of the argument (viz. {\tt n}) is easily inferable. In this
+case, since {\tt example} is of type {\tt ℕ → ℕ}, the argument {\tt n} has
+to be of type {\tt ℕ}, therefore leaving its type annotation unnecessary.
 
-This leaves the STLC rather small and limited but has the benefit of removing
-all \textit{infinite} calculations from the system \cite{}. Thus, the STLC is
-not \textit{Turing complete} and is thus the appropriate foundation for
-deriving a type theoretic system. This is because the STLC lends itself quite
-seamlessly as component of a larger, more expressive system. This flexibility
-is what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
-with more expressive types through simple extentions on its grammar for simple
-types.
-
-%% as the only form of dependency
-%% present in the STLC is (1), terms depending on terms (i.e. {\em term level}
-%% \(\lambda\)-abstraction).
+The STLC has limited expressibilty but has the benefit of removing all
+\textit{infinite} calculations from the system \cite{} and is thus the
+appropriate foundation for deriving a type theoretic system. This is because
+the STLC lends itself quite seamlessly as component of a larger, more expressive
+system. This flexibility is what we exploit to derive \(\lambda C\). To do this,
+we first extend the STLC with more expressive types through simple extentions on
+its grammar for simple types.
 
@@
 seperate entities in that terms only exist on {\em term level} and types
-on {\em type level}. Thus, as we mentioned earlier, this puts a constraint
-on the system to only allow \(\lambda\)-abstraction to happen at term level
+on {\em type level}. We return to the idea of {\em levels} later. Thus, as we
+mentioned earlier, this puts a constraint on the system to only allow
+\(\lambda\)-abstraction and function application to happen at term level
 (viz. definition of simple type). For example, the \textit{constant function}
@@
 treated differently and are only related in terms of {\em inhabitance}. This
-relation is reason for the limited expressivity of the STLC and is also the
+relation is the reason for the limited expressivity of the STLC and is also the
 method in which the STLC can be extended into \(\lambda C\).
@@
 Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
-\(\lambda2\) as simply an extension of the STLC. This system is adds
+\(\lambda2\) simply as an extension of the STLC. This system is adds
 {\em polymorphic} terms to the STLC (viz. Dependency (2)). For example,
@@
 Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
-are types. These two types (viz {\tt A} and {\tt B}) are the corresponding
+are types. These two types (viz. {\tt A} and {\tt B}) are the corresponding
 types of the third and fourth arguments of {\tt polyConst}. To clarify, we can
 instantiate {\tt polyConst} with \(\mathbb{N}\) and \(\mathbb{B}\) resulting
-in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us
-
+in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us:
 $${\tt const} = {\tt polyConst(ℕ,𝔹)}.$$
@@
 It is, however, necessary to constrain this expressivity with a more expressive
-type (e.g. {\tt (x : A) → ...}). Types that appear in this fashion are
+type (e.g. {\tt (x : Set) → ...}). Types that appear in this fashion are
 called \(\Pi\)-types. Taking the grammar for simple types, we can add \(\Pi\)-types,
@@
 \begin{code}
-t || t → t || ∀ (x : Set) → t
+Prim || t → t || ∀ (x : Set) → t
 \end{code}
@@
 \begin{code}
-t || t → t || ∀ (x : Set) → t || Set
+Prim || t → t || ∀ (x : Set) → t || Set
 \end{code}
@@
 \end{center}
-
 The curious effect of doing this allows us to construct expressions of type
@@
 This is to insure that we cannot write type expressions that fail to instantiate
-{\tt List} (e.g. {\tt List → A} v. {\tt List A → A}).
+{\tt List}. For example, the type expression {\tt List → A} is incorrect since
+{\tt List} cannot appear in a type expression without first being instantiated.
+To correct this, we can write this type expression as: {\tt List A → A}.
 
@@
 other extension necessary to consider our system fully-dependently typed, which,
-similar to our treatment of \(\lambda\underline{\omega}\) is also deceptively
+similar to our treatment of \(\lambda\underline{\omega}\), is also deceptively
 simple.
@@
 \end{center}
-
 Here, we've defined two ways in which {\tt n isEven} for all natural numbers
@@
 \subsection*{Calculus of Constructions and Remarks}
-To conclude our discussion over dependent types, we present the final grammar for
-types after extending \(\lambda\omega\) with \(\lambda P\):
+To conclude our discussion of dependent types, we present the final grammar for
+types after extending \(\lambda\omega\) with \(\lambda P\), resulting in
+\(\lambda C\):
 \begin{center}
@@
 \begin{code}
-t || t → t || ∀ (x : t) → t || Set
+Prim || t → t || ∀ (x : t) → t || Set
 \end{code}
@@
 which corresponds to the STLC, the base system. From here, we can either move
-inwards, upwards, or rightwards, which corresponds to the extension to the
-other systems marked on the respective edges. Using only these three movements,
-we eventually reach \(\lambda C\), leaving it in the {\em most powerful}
-position on the cube.
+inwards, upwards, or rightwards, which corresponds to the extension to the other
+systems marked on the respective edges. Using these three movements, we
+eventually reach \(\lambda C\), leaving it in the {\em most powerful} position
+on the cube.
 
@@
 
-\subsection*{Principles}
+\subsection*{Recursion and Induction Principles}
 Recursion and induction are one of the powerful tools available in a
@@
 The first argument given to these principles is sometimes called the {\em motive}.
-This simply means that {\tt C} is the type from which we will return a element
+This simply means that {\tt C} is the type from which we will return an element
 from. For recursion principles, this means the given output of the function we
@@
 \end{center}
-
 Next, we introduce the type \(\top\), which contains exactly one element,
@@
 There are, however, some trivial yet interesting proofs that involve the two.
-The first of these we also introduce a special kind of proof: {\it uniqueness
+For the first of these, we also introduce a special kind of proof: {\it uniqueness
 of identity} (UIP) \cite{}. In this proof, we set out to determine and classify
@@
 uip⊤ : (x : ⊤) → x ≡ *
-uip⊤ x = ind⊤ (λ top → top ≡ *) (refl *) x
---uip⊤ * = refl *
+uip⊤ = ind⊤ (λ top → top ≡ *) (refl *)
 \end{code}
@@
 which essentially repeats the information already given by our type expression.
-The second element, {\tt refl *}, is a proof that our motive holds in the event
-{\tt top} is {\tt *}, which is of type {\tt * ≡ *}. The last element {\tt x} is
-the expression initially passed to our proof of type \(\top\). Equivalently, we
-could have also written our proof as the third line above. This form of proof
-uses pattern matching and essentially asserts that the only element in \(\top\)
-is {\tt *}, which greatly simplifies our proof. Both methods of proof are
-correct, but, in general, using pattern matching is only safe when the values of
-proof expressions are not dependent one another (i.e., their values don't change
-depending on which expression we apply pattern matching to).
+The second element, {\tt refl *}, is a proof that our motive holds in the case that
+{\tt top} is {\tt *} and is of type {\tt * ≡ *}.
+
+Alternatively, we could have written our proof as follows:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+uip⊤ * = refl *
+\end{code}
+\end{minipage}
+\end{center}
+This form of proof uses pattern matching and essentially asserts that the only
+element in \(\top\) is {\tt *}, which greatly simplifies our proof. Both methods
+of proof are correct, but, in general, using pattern matching is only safe when
+the values of proof expressions are not dependent one another (i.e., their
+values don't change depending on which expression we apply pattern matching to).
 
@@
 
-\subsection*{Extension to Homotopy Type Theory}
+\subsection*{Extension to Homotopy Type Theory (maybe)}
 

### 2016-07-25 paper.tex
blah

@@
 creates a union between them in that every expression of type {\tt A + B} is
-either {\tt inl a}, where {\tt a : A}, or {\tt inr b}, where {\tt b : B}.
+either {\tt (inl a)}, where {\tt a : A}, or {\tt (inr b)}, where {\tt b : B}.
 

### 2016-07-25 paper.tex
more
yo

@@
 
+We now can write a simple proof that shows the union of \(\top\) and \(\top\)
+is equivalent to the Boolean type, that is {\tt ⊤ + ⊤} = {\tt 𝔹}. Alternatively,
+since \(\top\) is occasionally written as 1 and {\tt 𝔹} as 2, we name our proof
+as follows:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+1+1=2 : ⊤ + ⊤ → 𝔹
+1+1=2 = ?
+\end{code}
+\end{minipage}
+\end{center}
+To do this, we must associate every possible term of type {\tt ⊤ + ⊤} to a
+given Boolean (i.e., True or False). This works out quite simply since the
+type {\tt ⊤ + ⊤} only has two possible terms, that is {\tt (inl *)} and
+{\tt (inr *)}. Thus, we can write the proof as follows:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+1+1=2 = rec+ 𝔹 (λ inl* → True) (λ inr* → False)
+\end{code}
+\end{minipage}
+\end{center}
+To clarify, this form of equivalence is called ``judmental equality''. We
+go over the three different forms of equivalence in the following section.
+
+%% x and Sigma
+
 \section{Example Types and Sample Proofs}
@@
 \end{center}
-%% proof: 1 + 1 = 2
 

### 2016-07-26 paper.tex
more work

--- 2016-07-25 paper.tex
@@
 \begin{center}
-\begin{minipage}{0.2\textwidth}
+\begin{minipage}{0.3\textwidth}
 \begin{code}
@@
 \begin{center}
-\begin{minipage}{0.5\textwidth}
+\begin{minipage}{0.6\textwidth}
 \begin{code}
@@
 and {\tt ind}, for each {\em primitive type}: \(\perp\), \(\top\), +,
-\(\times\), and \(\Sigma\). The symbols + and \(\times\) signify disjunction
-and conjunction and, thus, should not be confused with addition and
-multiplication. For the latter, we use the names {\tt plus} and {\tt times}.
+and \(\times\). The symbols + and \(\times\) signify disjunction and conjunction
+and, thus, should not be confused with addition and multiplication. For the
+latter, we use the names {\tt plus} and {\tt times}.
 
@@
 creates a union between them in that every expression of type {\tt A + B} is
-either {\tt inl a}, where {\tt a : A}, or {\tt inr b}, where {\tt b : B}.
+either {\tt (inl a)} or {\tt (inr b)}, where {\tt a : A} and {\tt b : B}.
 
-\section{Example Types and Sample Proofs}
+Before we can provide our next example, we must first introduce the
+{\em pair type} or {\em logical conjunction}. We also provide the corresponding
+recursion and induction principle for the {\tt ×} type:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+data _×_ (A B : Set) : Set where
+  _,_ : A → B → A × B
+
+rec× : {A B : Set} (C : Set) → (A → B → C) → A × B → C
+rec× C f (a , b) = f a b
+
+ind× : {A B : Set} (C : A × B → Set) →
+       (∀ (x : A) (y : B) → C (x , y)) →
+       (x : A × B) → C x
+ind× C f (a , b) = f a b
+\end{code}
+\end{minipage}
+\end{center}
+
+We can now write a simple proof that shows the union of \(\top\) and \(\top\)
+is equivalent to the Boolean type, that is {\tt (⊤ + ⊤) ↔ 𝔹}.
+Alternatively, since \(\top\) is occasionally written as 1 and {\tt 𝔹} as 2, we
+name our proof as follows:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+1+1=2 : (⊤ + ⊤) ↔ 𝔹
+1+1=2 = ?
+\end{code}
+\end{minipage}
+\end{center}
+To do this, we must associate every possible term of type {\tt ⊤ + ⊤} to a
+given Boolean. This works out quite simply since the type {\tt ⊤ + ⊤} only has
+two possible terms, {\tt (inl *)} and {\tt (inr *)}, which we can associate with
+the two terms, {\tt True} and {\tt False}, in the type {\tt 𝔹} and vice-versa.
+Thus, we can write the proof as follows:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+1+1=2 = <> (f , g)
+  where f = rec+ 𝔹 (λ inl* → True) (λ inr* → False)
+        g = rec𝔹 (⊤ +' ⊤) (inl *) (inr *)
+\end{code}
+\end{minipage}
+\end{center}
+For our purposes, we use the constructor {\tt <>} for constructing a term of
+the bijection type. For simplicity, we can define a bijection as a conjunction
+of two logical implications:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+data _↔_ (A B : Set) : Set where
+  <>_ : (A → B) × (B → A) → (A ↔ B)
+\end{code}
+\end{minipage}
+\end{center}
+To clarify, this form of equivalence is called ``judmental equality'' (i.e.,
+equality by logical bijection). We go over the three different forms of
+equivalence in the following section.
+
+\section{More Types and Sample Proofs}
 In this section, we present the type of Booleans, Natural numbers, Lists,

### 2016-07-27 paper.tex
more work

@@
 
-\section{Dependent Types}
+\section{The World of Dependent Types}
 In this section, we introduce the concept of dependent types and their role in
@@
 \(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
-over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). In a later system, we
-allow \(\Pi\)-types to be parameterized over terms (e.g. {\tt ∀ (x : t) → t}).
+over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}).
+
+%% In a later system, we
+%% allow \(\Pi\)-types to be parameterized over terms (e.g. {\tt ∀ (x : t) → t}).
 
@@
 systems marked on the respective edges. Using these three movements, we
-eventually reach \(\lambda C\), leaving it in the {\em most powerful} position
+eventually reach \(\lambda C\), leaving it in the most powerful position
 on the cube.
@@
 prop' {A} perp = ind⊥ A perp
+
+absurd : ⊥ → (1 ≡ 2)
+absurd = rec⊥ (1 ≡ 2) 
 \end{code}
@@
 which we describe later on in this section. For now, we simply say that the
-sole constructor for the identity type is {\it refl x}, for all x.
+sole constructor for the identity type is {\tt (refl x)}, which is a proof that
+{\it x ≡ x}.
 
@@
 which essentially repeats the information already given by our type expression.
-The second element, {\tt refl *}, is a proof that our motive holds in the case that
+The second element, {\tt (refl *)}, is a proof that our motive holds in the case that
 {\tt top} is {\tt *} and is of type {\tt * ≡ *}.
@@
 \end{center}
-This form of proof uses pattern matching and essentially asserts that the only
-element in \(\top\) is {\tt *}, which greatly simplifies our proof. Both methods
-of proof are correct, but, in general, using pattern matching is only safe when
-the values of proof expressions are not dependent one another (i.e., their
-values don't change depending on which expression we apply pattern matching to).
+This form of proof uses pattern matching and asserts that the only element in
+\(\top\) is {\tt *}, which greatly simplifies our proof. Both methods of proof
+are correct, but, in general, using pattern matching is only safe when the
+values of proof expressions are not dependent one another (i.e., their values
+don't change depending on which expression we apply pattern matching to). This
+topic is formally known as {\it dependent pattern-matching}, which is a bit
+out of the scope of this paper. Thus, for our purposes, we refrain from using
+pattern matching in our proofs and instead use recursion and induction
+principles.
 
@@
 
-\section{More Types and Sample Proofs}
-In this section, we present the type of Booleans, Natural numbers, Lists,
-and the identity type. To further understand these types, we also include
-a brief description of the different forms of equality, namely definitional,
-judgmental and propositional. In type theory, the specific differences
-between the three are generally misunderstood but are important in understanding
-basic type theoretic proof methods, which we also go into some detail in
-this section with several examples.
+\section{Equivalence and Sample Proofs}
+In this section, we present the type of Natural numbers, Lists, and the identity
+type. To further understand these types, we also include a brief description of
+the different forms of equality, namely definitional, judgmental and
+propositional. In type theory, the specific differences between the three are
+generally misunderstood but are important in understanding basic type theoretic
+proof methods, which we also go into some detail in this section with several
+examples.
 
@@
 %% talk about definitional equality, judgmental equality, prop equality
+The first form of equality that is generally more familiar than the others.
+This form of equality is called {\it definitional equality} \cite{}. Simply put,
+definitional equality is equality through definition, which requires no proof.
+For example, we know that for any two natural numbers {\tt m} and {\tt n},
+{\tt m `plus` n} is equal to the sum of {\tt m} and {\tt n} by definition of
+{\tt `plus`}.
 
-\subsection*{Booleans}
+We have already briefly mentoined the second form of equality, judgmental
+equality \cite{}, which is equality by logical bijection. For example, if we
+have the propositions {\tt A} and {\tt B}, and from either {\tt A} or {\tt B} we
+can derive {\tt B} and {\tt A} respectively, then we can say that {\tt A} and
+{\tt B} are judgmentally equivalent. A proof of judgmental equality requires a
+proof similar to the one we have presented in the previous section for the
+proposition {\tt (⊤ + ⊤) ↔ 𝔹}.
+
+The final form of equivalence is quite a bit more complex than the previous two
+and requires a special kind of proof. Propositional equality is equality through
+an identity type (e.g. {\tt x ≡ y}). Thus, we now formally introduce the
+definition of the identity type:
+%% say something that this is the HoTT version of identity
+%% possibly also point out that the discrepancy between the two
+%% -- "PROPOSITIONAL EQUALITY"
+%% -- For a proposition, C, on a proof of x ≡ y for
+%% -- any x and y, if we can show that it holds for
+%% -- (refl x), then C holds for all proofs, p : x ≡ y.
+%% -- "STRICT EQUALITY"
+%% -- For a proposition, C, on a reflexive equality (x ≡ x),
+%% -- for any x, if we can show that it holds for (refl x), then
+%% -- C holds for all other reflexive equalities p : x ≡ x.
 \begin{center}
@@
 \begin{code}
-data 𝔹 : Set where
-  True  : 𝔹
-  False : 𝔹
-
-rec𝔹 : (C : Set) → C → C → 𝔹 → C
-rec𝔹 C t f True = t
-rec𝔹 C t f False = f
-
-ind𝔹 : (C : 𝔹 → Set) → C True → C False → (x : 𝔹) → C x
-ind𝔹 C t f True = t
-ind𝔹 C t f False = f
+data _≡_ {A : Set} : (a b : A) → Set where
+  refl : (a : A) → (a ≡ a)
 \end{code}
@@
 \end{center}
+Out of all the types we have introduced, the identity type is the first type
+parameterized by terms (viz. {\tt a} and {\tt b}). Until now, we have only
+introduced types parameterized by other types. Types paramterized by terms are
+classified as an {\em indexed type} \cite{}. Essentially, if we think of types
+as shapes (i.e., we can think of them as circles containing points representing
+their terms), indexed types change their shape depending on which terms are
+passed to them. For example, it is not the case that {\tt y ≡ x} automatically
+comes from a proof of {\tt x ≡ y}. This requires a proof!
+
+Before we can prove this, however, we must first introduce the recursion and
+induction principles of the identity type. Unlike the other types we have
+previously introduced, the identity type has two induction principles, both of
+which have their own appropriate use case \cite{}:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+J : {A : Set} (C : {x y : A} → x ≡ y → Set) →
+    (∀ (x : A) → C (refl x)) →
+    {x y : A} (p : x ≡ y) → C p
+J C c (refl x) = c x
+
+K : {A : Set} (C : {x : A} → x ≡ x → Set) →
+    (∀ (x : A) → C (refl x)) →
+    {x : A} (p : x ≡ x) → C p
+K C c (refl x) = c x
+
+rec≡ : {A : Set} {x y : A} (C : A → Set) →
+       (p : x ≡ y) → C x → C y
+rec≡ {A} {x} {y} C = 
+  J (λ {x} {y} _ → C x → C y)
+    (λ x → λ z → z)
+    {x} {y} 
+\end{code}
+\end{minipage}
+\end{center}
+
+%% \subsection*{Booleans}
+%% \begin{center}
+%% \begin{minipage}{0.9\textwidth}
+%% \begin{code}
+%% data 𝔹 : Set where
+%%   True  : 𝔹
+%%   False : 𝔹
+
+%% rec𝔹 : (C : Set) → C → C → 𝔹 → C
+%% rec𝔹 C t f True = t
+%% rec𝔹 C t f False = f
+
+%% ind𝔹 : (C : 𝔹 → Set) → C True → C False → (x : 𝔹) → C x
+%% ind𝔹 C t f True = t
+%% ind𝔹 C t f False = f
+%% \end{code}
+%% \end{minipage}
+%% \end{center}
 
@@
 
-\subsection*{Identity Types}
-%% say something that this is the HoTT version of identity
-%% possibly also point out that the discrepancy between the two
-\begin{center}
-\begin{minipage}{0.9\textwidth}
-\begin{code}
-infix  4  _≡_
-data _≡_ {ℓ} {A : Set ℓ} : (a b : A) → Set ℓ where
-  refl : (a : A) → (a ≡ a)
-
--- "PROPOSITIONAL EQUALITY"
--- For a proposition, C, on a proof of x ≡ y for
--- any x and y, if we can show that it holds for
--- (refl x), then C holds for all proofs, p : x ≡ y.
-J : {A : Set} (C : {x y : A} → x ≡ y → Set) →
-    (∀ (x : A) → C (refl x)) →
-    {x y : A} (p : x ≡ y) → C p
-J C c (refl x) = c x
-
--- "STRICT EQUALITY"
--- For a proposition, C, on a reflexive equality (x ≡ x),
--- for any x, if we can show that it holds for (refl x), then
--- C holds for all other reflexive equalities p : x ≡ x.
-K : {A : Set} (C : {x : A} → x ≡ x → Set) →
-    (∀ (x : A) → C (refl x)) →
-    {x : A} (p : x ≡ x) → C p
-K C c (refl x) = c x
-
-rec≡ : {A : Set} {x y : A} (C : A → Set) →
-       (p : x ≡ y) → C x → C y
-rec≡ {A} {x} {y} C = 
-  J (λ {x} {y} _ → C x → C y)
-    (λ x → λ z → z)
-    {x} {y} 
-\end{code}
-\end{minipage}
-\end{center}
-
 \section{Discussion}

### 2016-08-02 paper.tex
more

@@
 To clarify, this form of equivalence is called ``judmental equality'' (i.e.,
-equality by logical bijection). We go over the three different forms of
+equality by logical bijection) \cite{}. We go over the three different forms of
 equivalence in the following section.
@@
 %% talk about definitional equality, judgmental equality, prop equality
-The first form of equality that is generally more familiar than the others.
-This form of equality is called {\it definitional equality} \cite{}. Simply put,
-definitional equality is equality through definition, which requires no proof.
-For example, we know that for any two natural numbers {\tt m} and {\tt n},
-{\tt m `plus` n} is equal to the sum of {\tt m} and {\tt n} by definition of
-{\tt `plus`}.
+The first form of equality is much simpler than the others. This form of
+equality is called {\it definitional equality} \cite{}. Simply put, definitional
+equality is equality through definition, which requires no proof. For example,
+we know that for any two natural numbers {\tt m} and {\tt n}, {\tt m `plus` n}
+is equal to the sum of {\tt m} and {\tt n} by definition of {\tt `plus`}.
 
-We have already briefly mentoined the second form of equality, judgmental
-equality \cite{}, which is equality by logical bijection. For example, if we
-have the propositions {\tt A} and {\tt B}, and from either {\tt A} or {\tt B} we
-can derive {\tt B} and {\tt A} respectively, then we can say that {\tt A} and
+We have already briefly mentioned the second form of equality, judgmental
+equality, which is equality by logical bijection. For example, if we have the
+propositions {\tt A} and {\tt B}, and from either {\tt A} or {\tt B} we can
+derive {\tt B} and {\tt A} respectively, then we can say that {\tt A} and
 {\tt B} are judgmentally equivalent. A proof of judgmental equality requires a
@@
 parameterized by terms (viz. {\tt a} and {\tt b}). Until now, we have only
-introduced types parameterized by other types. Types paramterized by terms are
-classified as an {\em indexed type} \cite{}. Essentially, if we think of types
+introduced types parameterized by other types. Types parameterized by terms are
+classified as an {\em indexed types} \cite{}. Essentially, if we think of types
 as shapes (i.e., we can think of them as circles containing points representing
@@
 passed to them. For example, it is not the case that {\tt y ≡ x} automatically
-comes from a proof of {\tt x ≡ y}. This requires a proof!
+comes from a proof of {\tt x ≡ y}. This requires a proof of the symmetry of
+identity types.
 
@@
     {x} {y} 
+\end{code}
+\end{minipage}
+\end{center}
+The two induction principles of the identity type (viz. {\tt J} and {\tt K})
+refer to the two ways in which an identity type can be used in a proof, where
+{\tt J} is more general. This simply means that {\tt J} can be used for a wider
+variety of proofs concerning identity types.
+
+At first glance, the difference between the two induction principles is clear
+when looking at the {\tt C} parameter in both. In {\tt J}, {\tt C} is
+parameterized with two possibly distinct terms (viz. {\tt x} and {\tt y}),
+whereas {\tt K} only mentions one, {\tt x}. This distinction is what allows
+{\tt J} to be used for {\em special} identity proofs, that is, in proofs
+for identity types of more complex forms of equality (i.e., propositional
+equality).
+
+As an example, let's prove the symmetry of identity types using {\tt J}:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
+sym_J = J ? ?
+\end{code}
+\end{minipage}
+\end{center}
+Now, we must fill in both instances of {\tt ?} in the above proof. Similar to
+previous proofs, the first of these is our motive and is simply a restatement of
+our type statement. In these forms of proofs, we commonly annotate parameterized
+identity types with their given equality (viz. {\tt x=y}):
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
+sym_J = J (λ {x y} x=y → y ≡ x) ?
+\end{code}
+\end{minipage}
+\end{center}
+Now, the last {\tt ?} must be filled in with a proof that our motive holds in
+the case that {\tt y} is {\tt x}, that is, we must prove that
+{\tt (x : A) → x ≡ x}, which is the type of {\tt refl}. Thus, our completed
+proof is the following:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
+sym_J = J (λ {x y} x=y → y ≡ x) refl
 \end{code}

### 2016-08-03 paper.tex
getting there

@@
 
-\section{Equivalence and Sample Proofs}
+\section{Equivalence and Other Types}
 In this section, we present the type of Natural numbers, Lists, and the identity
@@
 
-\subsection*{Equality}
+\subsection*{Equality and Identity Types}
 %% talk about definitional equality, judgmental equality, prop equality
@@
 an identity type (e.g. {\tt x ≡ y}). Thus, we now formally introduce the
-definition of the identity type:
+definition of the identity type.
 %% say something that this is the HoTT version of identity
@@
 Before we can prove this, however, we must first introduce the recursion and
-induction principles of the identity type. Unlike the other types we have
-previously introduced, the identity type has two induction principles, both of
-which have their own appropriate use case \cite{}:
+induction principles of the identity type:
 \begin{center}
@@
 J C c (refl x) = c x
-
-K : {A : Set} (C : {x : A} → x ≡ x → Set) →
-    (∀ (x : A) → C (refl x)) →
-    {x : A} (p : x ≡ x) → C p
-K C c (refl x) = c x
 
@@
 \end{center}
-The two induction principles of the identity type (viz. {\tt J} and {\tt K})
-refer to the two ways in which an identity type can be used in a proof, where
-{\tt J} is more general. This simply means that {\tt J} can be used for a wider
-variety of proofs concerning identity types.
+The more general induction principle for identity types is called {\tt J}.
+Later, we discuss a simpler, yet more restricted version. For now, we take
+the time to introduce the more general version which is applicable to every
+possible identity type \cite{}.
 
-At first glance, the difference between the two induction principles is clear
-when looking at the {\tt C} parameter in both. In {\tt J}, {\tt C} is
-parameterized with two possibly distinct terms (viz. {\tt x} and {\tt y}),
-whereas {\tt K} only mentions one, {\tt x}. This distinction is what allows
-{\tt J} to be used for {\em special} identity proofs, that is, in proofs
-for identity types of more complex forms of equality (i.e., propositional
-equality).
+Since identity types are parameterized over two terms, naturally, the motive
+parameter (viz. {\tt C}) of their induction principle is also parameterized
+by terms (viz. {\tt x} and {\tt y}). The simple reading of {\tt J} is read
+similar to other induction principles: given a proposition, {\tt C}, on an
+arbitrary identity type, if we can prove that {\tt C} holds for {\tt refl x},
+for all {\tt x}, then {\tt C} holds for all identity types. This becomes a bit
+more intuitive when we think of what a proof of equality entails: if we want
+to prove something for a proof of equality, it suffices to prove it for the
+simplest form of equality (i.e., a reflexive equality, {\tt refl}). One can
+also note the similarities between {\tt J} and the induction principle for
+{\tt ⊤}, since these types both have a single constructor (i.e., {\tt refl}
+and {\tt *}, respectively).
 
-As an example, let's prove the symmetry of identity types using {\tt J}:
+Now, let's prove the symmetry of identity types using {\tt J}:
 \begin{center}
@@
 \begin{code}
-sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
-sym_J = J ? ?
+sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
+sym = J ? ?
 \end{code}
@@
 \end{center}
-Now, we must fill in both instances of {\tt ?} in the above proof. Similar to
+Next, we must fill in both instances of {\tt ?} in the above proof. Similar to
 previous proofs, the first of these is our motive and is simply a restatement of
@@
 \begin{code}
-sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
-sym_J = J (λ {x y} x=y → y ≡ x) ?
+sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
+sym = J (λ {x y} x=y → y ≡ x) ?
 \end{code}
@@
 \begin{code}
-sym_J : {A : Set} {x y : A} → x ≡ y → y ≡ x
-sym_J = J (λ {x y} x=y → y ≡ x) refl
+sym : {A : Set} {x y : A} → x ≡ y → y ≡ x
+sym = J (λ {x y} x=y → y ≡ x) refl
 \end{code}
@@
 \end{center}
+To reiterate, this is an example of propositional equality. If this were a
+definitional or judgmental equality, we would not need to prove symmetry, since
+these two forms of equality are automatically reflexive, symmetric, associative,
+and transitive. This is the primary and most important difference between other
+forms of equality and propositional equality.
 
-%% \subsection*{Booleans}
-%% \begin{center}
-%% \begin{minipage}{0.9\textwidth}
-%% \begin{code}
-%% data 𝔹 : Set where
-%%   True  : 𝔹
-%%   False : 𝔹
+A curious fact about {\tt J} is that it does not lend itself to a uniqueness of
+identity proof, UIP. In fact, for many reasons, this is actually beneficial and
+allows the identity type to be used outside of simple proofs of equality \cite{}.
+We discuss this in the final section of this paper and compare {\tt J}
+side-by-side with another, less general induction principle for identity types.
 
-%% rec𝔹 : (C : Set) → C → C → 𝔹 → C
-%% rec𝔹 C t f True = t
-%% rec𝔹 C t f False = f
-
-%% ind𝔹 : (C : 𝔹 → Set) → C True → C False → (x : 𝔹) → C x
-%% ind𝔹 C t f True = t
-%% ind𝔹 C t f False = f
-%% \end{code}
-%% \end{minipage}
-%% \end{center}
+\subsection*{Other types}
+We now simply provide the remaining types for Natural numbers and lists. With
+what we have discussed, these types should not require much explanation.
 
@@
 
-\section{Discussion}
-Say a few things about the paper.
+\section{Exercises}
+We have now seen and discussed all the types necessary to start proving more
+complex proofs. As an exercise to the reader, we include a list of propositions
+to prove. We invite the reader to use only the appropriate induction and
+recursion principles in each proof. It helps to first attempt to prove these
+propositions using simple pattern matching. As a hint, use a recursion principle
+when the proof behaves like a simple function (i.e., defining {\tt `plus`}) and
+induction principle when proving propositions {\tt C} for all terms of a given
+type (i.e., like {\tt sym}).
+
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+-- Identity Types
+congruence : {A : Set} {x y : A} (P : A → Set) →
+             (p : x ≡ y) → P x → P y
+transitivity : {A : Set} {x y z : A} → (x ≡ y) →
+               (y ≡ z) → (x ≡ z)
+
+-- Natural numbers (requires the above)
+plus : ℕ → ℕ → ℕ
+plus-left-unit : (i : ℕ) → plus 0 i ≡ i
+plus-right-unit : (i : ℕ) → plus i 0 ≡ i
+plus-successor : (i j : ℕ) → plus (suc i) j ≡ plus i (suc j)
+plus-commutativity : (i j : ℕ) → plus i j ≡ plus j i
+plus-associativity : (x y z : ℕ) →
+                     plus x (plus y z) ≡ plus (plus x y) z
+
+-- Lists
+_++_ : {A : Set} → List A → List A → List A
+reverse : {A : Set} → List A → List A
+append-left-unit : {A : Set} (xs : List A) → Nil ++ xs ≡ xs
+append-right-unit : {A : Set} (xs : List A) → xs ++ Nil ≡ xs
+append-associativity : {A : Set} (xs ys zs : List A) →
+                       xs ++ (ys ++ zs) ≡ (xs ++ ys) ++ zs
+append-distributivity : {A : Set} (xs ys : List A) →
+                        reverse (xs ++ ys) ≡ (reverse ys) ++ (reverse xs) 
+double-reverse : {A : Set} (xs : List A) → reverse (reverse xs) ≡ xs
+
+\end{code}
+\end{minipage}
+\end{center}
 

### 2016-08-05 paper.tex
polishing

@@
 theory have a viable application in mathematics but also a growing interest
-in the field of computer science. In both fields, type theory provides a richer
-world of types and expressibiilty.
+in the field of computer science where types find themselves in great use in
+many popular programming languages. In both fields, type theory provides a
+richer world of types and expressibiilty.
 
 The study of type theory goes hand in hand with the study of dependent types,
-which is why we first take the time to do familiarize ourselves with them
-(Section 2). We then continue with the foundational concepts of type theory of
-Per Martin-L\"of \cite{}. For this, we recommend using a proof assistant, such
-as Agda. In our examples, we refer to Agda syntax rules.
+which is why we first take the time to familiarize ourselves with them
+(Section 2). We then continue with the fundamental concepts of type theory of
+Per Martin-L\"of \cite{}. For this, we recommend the reader have a running proof
+assistant, such as Agda. In our examples, we refer to Agda syntax rules, and
+using a proof assistant should help develop the necessary thought processes to
+successfully do type theoretic proofs.
 
@@
 (\(\lambda C\)) \cite{}. Here, the word \textit{fully} signifies that \(\lambda C\)
-features all four forms of dependency mentioned above.
+includes all four forms of dependency mentioned above.
 
@@
 
-The STLC has limited expressibilty but has the benefit of removing all
-\textit{infinite} calculations from the system \cite{} and is thus the
-appropriate foundation for deriving a type theoretic system. This is because
-the STLC lends itself quite seamlessly as component of a larger, more expressive
-system. This flexibility is what we exploit to derive \(\lambda C\). To do this,
-we first extend the STLC with more expressive types through simple extentions on
-its grammar for simple types.
+The STLC has limited expressibilty but has the benefit of no longer having any
+infinite calculations \cite{} and is thus the appropriate foundation for
+deriving a type theoretic system. This is because the STLC lends itself quite
+seamlessly as component of larger, more expressive systems. This flexibility is
+what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
+with more expressive types through simple extentions of its grammar of simple
+types.
 
@@
 Given that {\tt const} is of type {\tt ℕ → 𝔹 → ℕ}, the term that inhabits this
-type is {\tt λ n b → n}. In this example, we can see that terms and types are
-treated differently and are only related in terms of {\em inhabitance}. This
-relation is the reason for the limited expressivity of the STLC and is also the
-method in which the STLC can be extended into \(\lambda C\).
+type is {\tt λ n b → n}, where {\tt n} and {\tt b} are typed {\tt ℕ} and {\tt 𝔹},
+respectively. Here, we are also limited by the fact that this constant function can
+only be applied to natural numbers and booleans. It is not the most general
+(i.e., applicable to any two terms) way to define such a function. Thus, in this
+example, we can see that terms and types are treated differently and are only
+related in terms of {\em inhabitance}. This relation is the reason for the
+limited expressivity of the STLC and is also the method in which the STLC can be
+extended into a more expressive system.
 
@@
 Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
-\(\lambda2\) simply as an extension of the STLC. This system is adds
-{\em polymorphic} terms to the STLC (viz. Dependency (2)). For example,
-let's take the definition of {\tt const} above and change it to define
-{\tt polyConst}:
+\(\lambda2\) as an extension of the STLC. This system is adds {\em polymorphic}
+terms to the STLC (Dependency (2)). For example, let's take the definition of
+{\tt const} above and change it to define {\tt polyConst}:
 \begin{center}
@@
 polyConst A B = λ a b → a
--- Works for any type A and B
 \end{code}
@@
 \end{center}
-
 Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
@@
 type (e.g. {\tt (x : Set) → ...}). Types that appear in this fashion are
-called \(\Pi\)-types. Taking the grammar for simple types, we can add \(\Pi\)-types,
-resulting in the grammar for types ({\tt t}) in \(\lambda2\):
+called \(\Pi\)-types. Extending the grammar for simple types with \(\Pi\)-types
+results in the grammar for types ({\tt t}) in \(\lambda2\):
 
@@
 \(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
-over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}).
-
-%% In a later system, we
-%% allow \(\Pi\)-types to be parameterized over terms (e.g. {\tt ∀ (x : t) → t}).
+over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). Before we can used
+\(\Pi\)-types to their full potential, we must first introduce other forms of
+type and term dependency.
 
@@
 The curious effect of doing this allows us to construct expressions of type
-{\tt t → t}, where either {\tt t} can have subexpressions containing {\tt Set}.
+\mbox{{\tt t → t}}, where either {\tt t} can have subexpressions containing {\tt Set}.
 Let's take the familiar definition of the list data structure:
@@
 {\tt List} cannot appear in a type expression without first being instantiated.
-To correct this, we can write this type expression as: {\tt List A → A}.
+To correct this, we can write this type expression as {\tt List A → A}.
 
@@
 system of \(\lambda P\), allows us to construct {\em type families} which are
-types themselves and are meant to directly correspond to propositions in
+themselves types and are meant to directly correspond to propositions in
 higher-order logic.
@@
 \end{center}
-Here, we've defined two ways in which {\tt n isEven} for all natural numbers
-({\tt n}) can be proven. Returning to the idea of proof-relevant mathematics,
+Here, we've defined two ways in which {\tt n isEven} for all natural numbers,
+{\tt n}, can be proven. Returning to the idea of proof-relevant mathematics,
 we do this by use of its two type constructors {\tt ZEven} and {\tt NEven},
@@
 proof that 0 is an even number. {\tt NEven} then defines the property that if
-a given natural number ({\tt n}) is even (viz. {\tt n isEven}), then {\tt n + 2}
+a given natural number, {\tt n}, is even (viz. {\tt n isEven}), then {\tt n + 2}
 is even (viz. {\tt (Suc (Suc n)) isEven}). To clarify how this type can be used,
@@
 While it might seem intuitive and more effective to instead try to prove that 5
-is {\em not} even, this, however, also doesn't work. This is due to the
+is {\em not} even, this, however, also does not work. This is due to the
 treatment of negation in Intuitionistic Logic. To prove that 5 is not even is
@@
 %% Curry-Howard?
-In this section, we introduce the foundational concepts of Martin-L\"of Type
+In this section, we introduce the fundamental concepts of Martin-L\"of Type
 Theory \cite{}. We discuss recursion, induction, and as well several primitive type
@@
 (viz. {\tt T} and {\tt (t : T)}). The only difference is that the induction
-principle introduces a type family, as opposed to a non-parameterized type in the
-recursion principle. We discuss this in Section 4.
+principle introduces a type family (e.g. similar to {\tt isEven}), while a
+recursion principle does not.
 
@@
 \end{center}
+For clarity, we annotate terms with their respective type (i.e., {\tt ¬¬¬a} is
+of type {\tt ¬¬¬A}). We do this quite often when writing proofs.
+
 Next, we introduce the type \(\top\), which contains exactly one element,
@@
 This second expression, {\it x ≡} {\tt *}, is an example of an {\it identity type},
-which we describe later on in this section. For now, we simply say that the
+which we describe later on (Section 4). For now, we simply say that the
 sole constructor for the identity type is {\tt (refl x)}, which is a proof that
-{\it x ≡ x}.
+{\it x ≡ x}, for any term {\tt x}.
 
@@
 Here, we have taken our motive to be the expression {\tt (λ top → top ≡ *)},
-which essentially repeats the information already given by our type expression.
-The second element, {\tt (refl *)}, is a proof that our motive holds in the case that
-{\tt top} is {\tt *} and is of type {\tt * ≡ *}.
+which essentially repeats the information already given by our type expression
+(viz. {\tt (x : ⊤) → x ≡ *}). The second element, {\tt (refl *)}, is a proof
+that our motive holds in the case that {\tt top} is {\tt *} and is of type
+{\tt * ≡ *}.
 
@@
 creates a union between them in that every expression of type {\tt A + B} is
-either {\tt (inl a)} or {\tt (inr b)}, where {\tt a : A} and {\tt b : B}.
+either {\tt (inl a)} or {\tt (inr b)}, where {\tt a} and {\tt b} are typed
+{\tt A} and {\tt B}, respectively.
 
@@
   where f = rec+ 𝔹 (λ inl* → True) (λ inr* → False)
-        g = rec𝔹 (⊤ +' ⊤) (inl *) (inr *)
+        g = rec𝔹 (⊤ + ⊤) (inl *) (inr *)
 \end{code}
@@
 equality is equality through definition, which requires no proof. For example,
-we know that for any two natural numbers {\tt m} and {\tt n}, {\tt m `plus` n}
-is equal to the sum of {\tt m} and {\tt n} by definition of {\tt `plus`}.
+we know that for the natural numbers, {\tt 1 `plus` 1} is equal to {\tt 2} by
+the definition of {\tt plus}.
 
@@
 previous proofs, the first of these is our motive and is simply a restatement of
-our type statement. In these forms of proofs, we commonly annotate parameterized
-identity types with their given equality (viz. {\tt x=y}):
+our type statement. As we have done earlier, we annotate parameterized identity
+types with their given equality (viz. {\tt x=y}):
 \begin{center}
@@
 \subsection*{Natural Numbers}
+Natural numbers have two constructors: {\tt 0} and {\tt suc}. The latter of
+these is a function that simply adds 1 to any natural number. Natural numbers
+are the simplest example of an infinite type (i.e., it encodes every positive
+whole number, {\tt n}, such that {\tt n} greater than or equal to {\tt 0}).
+
 \begin{center}
@@
 \subsection*{List/Orderings}
+Lists are similar to natural numbers in that their type definition also has two
+constructors: {\tt Nil}, which represents the empty list, and a function that
+adds an element to the front of a given list, commonly read as {\em cons}.
+
 \begin{center}
@@
 -- Natural numbers (requires the above)
-plus : ℕ → ℕ → ℕ
-plus-left-unit : (i : ℕ) → plus 0 i ≡ i
-plus-right-unit : (i : ℕ) → plus i 0 ≡ i
-plus-successor : (i j : ℕ) → plus (suc i) j ≡ plus i (suc j)
-plus-commutativity : (i j : ℕ) → plus i j ≡ plus j i
+_plus_ : ℕ → ℕ → ℕ
+plus-left-unit : (i : ℕ) → 0 plus i ≡ i
+plus-right-unit : (i : ℕ) → i plus 0 ≡ i
+plus-successor : (i j : ℕ) → (suc i) plus j ≡ i plus (suc j)
+plus-commutativity : (i j : ℕ) → i plus j ≡ j plus i
 plus-associativity : (x y z : ℕ) →
-                     plus x (plus y z) ≡ plus (plus x y) z
+                     x plus (y plus z) ≡ (x plus y) plus z
 
@@
 \subsection*{Extension to Homotopy Type Theory (maybe)}
+\subsection*{Conclusion and Acknowledgments}
 

### 2016-08-05 paper.tex
more

@@
 induction principle when proving propositions {\tt C} for all terms of a given
-type (i.e., like {\tt sym}).
+type (i.e., like {\tt sym}). As an example, the first proof of each set of
+proofs is provided:
 
@@
              (p : x ≡ y) → P x → P y
+congruence P = 
+  J (λ {x y} x=y → P x → P y)
+    (λ x px → px)
+
 transitivity : {A : Set} {x y z : A} → (x ≡ y) →
                (y ≡ z) → (x ≡ z)
+\end{code}
+\end{minipage}
+\end{center}
 
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
 -- Natural numbers (requires the above)
 _plus_ : ℕ → ℕ → ℕ
+m plus n = f m n where
+  f = recℕ (ℕ → ℕ) (λ n → n) (λ m r n → suc (r n))
+
 plus-left-unit : (i : ℕ) → 0 plus i ≡ i
+plus-left-unit = 
+  indℕ (λ i → 0 plus i ≡ i)
+        (refl zero)
+        (λ x _ → refl (suc x))
+
 plus-right-unit : (i : ℕ) → i plus 0 ≡ i
@@
                      x plus (y plus z) ≡ (x plus y) plus z
+\end{code}
+\end{minipage}
+\end{center}
 
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
 -- Lists
 _++_ : {A : Set} → List A → List A → List A
+_++_ = append where
+  append : {A : Set} → List A → List A → List A
+  append {A} = recList (List A → List A)
+                       (λ ys → ys)
+                       (λ a xs f ys → a :: (f ys))
+
 reverse : {A : Set} → List A → List A
@@
 double-reverse : {A : Set} (xs : List A) → reverse (reverse xs) ≡ xs
-
 \end{code}

### 2016-08-09 paper.tex
worked hott part

@@
 \section{Introduction}
-We present an introduction to dependent types and (Martin-L\"of) type theory.
+We present an overview of dependent types and (Martin-L\"of) type theory.
 At its core, type theory seeks to serve as a possible foundation of mathematics,
@@
 which is why we first take the time to familiarize ourselves with them
-(Section 2). We then continue with the fundamental concepts of type theory of
+(Section 2). We then continue with the fundamental concepts of the type theory of
 Per Martin-L\"of \cite{}. For this, we recommend the reader have a running proof
@@
 is drawn in constrast with the term {\em simple type}, where the STLC derives
-its namesake. A simple type ({\tt t}) is either a {\em primitive type}, {\tt Prim},
+its namesake. A simple type, {\tt t}, is either a {\em primitive type}, {\tt Prim},
 which is the set of all types ``pre-defined'' (e.g. lists, numbers, booleans),
@@
 infinite calculations \cite{} and is thus the appropriate foundation for
-deriving a type theoretic system. This is because the STLC lends itself quite
-seamlessly as component of larger, more expressive systems. This flexibility is
+deriving a type theoretic system. On top of this, the STLC lends itself quite
+seamlessly as a component of larger, more expressive systems. This flexibility is
 what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
-with more expressive types through simple extentions of its grammar of simple
-types.
-
-The key difference between the STLC and \(\lambda C\) are the places in which
-terms and types are allowed to be. In the STLC, terms and types are entirely
-seperate entities in that terms only exist on {\em term level} and types
-on {\em type level}. We return to the idea of {\em levels} later. Thus, as we
-mentioned earlier, this puts a constraint on the system to only allow
-\(\lambda\)-abstraction and function application to happen at term level
-(viz. definition of simple type). For example, the \textit{constant function}
-over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC is defined:
+with more expressive types through simple extentions of its type grammar.
+
+With this, the key difference between the STLC and \(\lambda C\) are the places
+in which terms and types are allowed to be within a given expression. In the
+STLC, terms and types are entirely seperate entities in that terms only exist on
+{\em term level} and types on {\em type level}. We return to the idea of
+{\em levels} later. Thus, as we mentioned earlier, this puts a constraint on the
+system to only allow \(\lambda\)-abstraction and function application to happen
+at term level (viz. definition of simple type). For example, the
+{\em constant function} over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC
+is defined:
 \begin{center}
@@
 in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us:
-$${\tt const} = {\tt polyConst(ℕ,𝔹)}.$$
+$${\tt polyConst(ℕ,𝔹)} = {\tt const}.$$
 Thus, adding polymorphism to the STLC has the effect of extending the notion
@@
 called \(\Pi\)-types. Extending the grammar for simple types with \(\Pi\)-types
-results in the grammar for types ({\tt t}) in \(\lambda2\):
+results in the type grammar of \(\lambda2\):
 
@@
 \(\Pi\)-types are actually significantly more expressive than the ones in
-\(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
-over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). Before we can used
+\(\lambda2\). In \(\lambda2\), they are only allowed to be parameterized
+over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). Before we can use
 \(\Pi\)-types to their full potential, we must first introduce other forms of
@@
 
-Thus, the system we are introducing is \(\lambda\omega\), which, unlike
+Thus, the system we are introducing here is \(\lambda\omega\), which, unlike
 \(\lambda\underline{\omega}\), features a more diverse grammar of types (i.e.,
@@
 distinction for the first two levels: 42 is NOT of type {\tt Set}, while
-\(\mathbb{N}\) is both of type {\tt Set}, {\tt Set₁} and onwards. Terms are
+\(\mathbb{N}\) is of type {\tt Set}, {\tt Set₁} and so on. Terms are
 therefore treated uniquely, since they only exist as elements of types.
 
-We can now revise the definition of types ({\tt t}) to include all instances of
+We can now revise the definition of types to include all instances of
 {\tt Set} in our infinite universe. We notate this as {\tt Set} below:
@@
 themselves types and are meant to directly correspond to propositions in
-higher-order logic.
+higher-order logic \cite{}.
 
@@
 over natural numbers. When writing propositions, we write them as their equivalent
-type definition (similar to {\tt List}). Consequently, the proposition
-{\tt isEven} is an example of a dependent type, since its {\em meaning} depends
-on the natural number passed to it (types depending on terms).
+type definition (similar to how {\tt List} is defined).
+%% Consequently, the proposition
+%% {\tt isEven} is an example of a dependent type, since its {\em meaning} depends
+%% on the natural number passed to it (types depending on terms).
 
@@
 This makes sense since we use types to correspond to logical propositions
-(e.g. {\tt isEven}). ``Falsity'' then naturally corresponds to the un-inhabitability
-of a given type. This can be seen if we try to prove that 5 is even using
-{\tt isEven}. The proof seems to get ``stuck'' at trying to prove that 1 is even.
+(e.g. {\tt isEven}) and types are either inhabitted or un-inhabitted. Naturally,
+``Falsity'' then corresponds to the un-inhabitability of a given type. This can
+be seen if we try to prove that 5 is even using {\tt isEven}. The proof seems to
+get ``stuck'' at trying to prove that 1 is even.
 
@@
 \subsection*{Extension to Homotopy Type Theory (maybe)}
+Earlier, we briefly discussed how types can be viewed as shapes, with their
+respective terms consequently viewable as points contained with shapes. This
+concept is important in understanding another way to view the concepts of type
+theory: {\em homotopy type theory} (HoTT).
+
+In fact, a lot of what HoTT seeks to do stems directly from a single concept in
+regular type theory: the induction principle of identity types! This is why we
+briefly discussed identity types having {\em another}, less general induction
+principle. This less general induction principle is known as {\tt K} and is
+defined as follows:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+K : {A : Set} (C : {x : A} → x ≡ x → Set) →
+    (∀ (x : A) → C (refl x)) →
+    {x : A} (p : x ≡ x) → C p
+K C c (refl x) = c x
+\end{code}
+\end{minipage}
+\end{center}
+{\tt K} is similar to {\tt J} but asserts that the only proof of equality is
+{\tt refl} (i.e., the only equality is reflexive equality). In fact, we can
+prove this using {\tt K}:
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+UIP≡ : {A : Set} {x : A} (p : x ≡ x) → p ≡ (refl x)
+UIP≡ = 
+  K (λ {x} x=x → x=x ≡ (refl x))
+    (λ x → refl (refl x))
+\end{code}
+\end{minipage}
+\end{center}
+As we mentioned earlier, we cannot prove UIP for identity types using {\tt J},
+since the identity type passed to {\tt J} is parameterized over two possibly
+different terms, while {\tt K} only applies to reflexive proofs of equality.
+In some cases, {\tt K} might be all that one would need to do a given number
+of proofs concerning identity types, however, it limits the overall application
+and expressivity of identity types.
 \subsection*{Conclusion and Acknowledgments}

### 2016-08-10 paper.tex
forgot to save
more

@@
 
+ap : {A : Set} {B : Set} (f : A → B) →
+     {x y : A} → (x ≡ y) → (f x ≡ f y)
 transitivity : {A : Set} {x y z : A} → (x ≡ y) →
@@
 
-reverse : {A : Set} → List A → List A
+rev : {A : Set} → List A → List A
 append-left-unit : {A : Set} (xs : List A) → Nil ++ xs ≡ xs
@@
 append-distributivity : {A : Set} (xs ys : List A) →
-                        reverse (xs ++ ys) ≡ (reverse ys) ++ (reverse xs) 
-double-reverse : {A : Set} (xs : List A) → reverse (reverse xs) ≡ xs
-\end{code}
-\end{minipage}
-\end{center}
-
-\subsection*{Extension to Homotopy Type Theory (maybe)}
+                        rev (xs ++ ys) ≡ (rev ys) ++ (rev xs) 
+double-rev : {A : Set} (xs : List A) → rev (rev xs) ≡ xs
+\end{code}
+\end{minipage}
+\end{center}
+
+\subsection*{A little more on Identity Types}
 Earlier, we briefly discussed how types can be viewed as shapes, with their
-respective terms consequently viewable as points contained with shapes. This
-concept is important in understanding another way to view the concepts of type
-theory: {\em homotopy type theory} (HoTT).
+respective terms viewable as points contained within them. This view of types
+is important in understanding another way to apply type theory:
+{\em homotopy type theory} (HoTT).
 
@@
 {\tt refl} (i.e., the only equality is reflexive equality). In fact, we can
-prove this using {\tt K}:
+this concept using {\tt K}:
 \begin{center}
@@
 \end{center}
-As we mentioned earlier, we cannot prove UIP for identity types using {\tt J},
-since the identity type passed to {\tt J} is parameterized over two possibly
-different terms, while {\tt K} only applies to reflexive proofs of equality.
-In some cases, {\tt K} might be all that one would need to do a given number
-of proofs concerning identity types, however, it limits the overall application
-and expressivity of identity types.
+As we mentioned earlier, we cannot prove {\tt UIP≡} using {\tt J}, since the
+identity type passed to {\tt J} is parameterized over two possibly different
+terms, while {\tt K} only applies to reflexive proofs of equality. Another way
+to think of this is that {\tt J} applies to all two terms that are proven equal
+{\em in some way}, while {\tt K} can only be used for terms that either
+reducible to each other or are, in some way, the same. In some cases, {\tt K}
+might be all that one would need to do a given number of proofs concerning
+identity types, however, it limits the overall application and expressivity of
+identity types.
+
+In HoTT, identity types are interpretted rather differently than in regular type
+theory. In HoTT, identity types correspond to {\em topological paths} that
+connect two elements. Here, we say {\em elements} because these paths can
+connect points, paths, paths of paths, etc. Another way to say this is that HoTT
+has the ability to reason about the {\em equality of equalities}. This is the
+reason for complexity of {\tt J} and is also the reason why {\tt K} is not
+applicable to HoTT.
+
+To end our brief discussion on identity types, it helps to further to solidify
+the concept of how identity types can be interpretted as topological paths in
+HoTT. For simplicity, we show how all the methods in which we can interact with
+an identity type correspond to a way one could think of interacting with a path.
+\begin{center}
+\begin{minipage}{0.9\textwidth}
+\begin{code}
+Identity Type → Path
+
+J → path induction
+symmetry → path inverse
+ap → function application on paths
+congruence → proposition interaction on paths
+transitiviy → path concatenation
+\end{code}
+\end{minipage}
+\end{center}
+
 \subsection*{Conclusion and Acknowledgments}
-
+At its core, the main purpose of type theory is to propose a new way of doing
+mathematics. The main benefit of choosing to do so stems from the close
+connection between types, constructive logic and the field of computer science.
+Thus, through simple act of type checking, which was made possible due in part
+to type theory being founded on the STLC, proofs and programs can now be
+verified and corrected with the help of a computer equipped with a proof
+assistant, which has given both mathematics and computer science the powerful
+concept of {\em proof verification}.
+
+I would like to thank Daniel P. Friedman for guiding me as I explored the world
+of types and type theory. 
 \end{document}
